---
title: "Merck - CS-CORE PBMC 3k"
author: "Noor Sohail"
output:
   html_document:
      code_folding: hide
      df_print: paged
      highlights: pygments
      number_sections: true
      self_contained: true
      theme: default
      toc: true
      toc_float:
         collapsed: true
         smooth_scroll: true
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
# Turn off Warnings and other console output messages from the whole document
```

```{r}
#' Create sub-chunks for plots
#' Plotting function to generate plots of a specific dimension when html is knitted
#'
#' taken from: https://stackoverflow.com/questions/15365829/dynamic-height-and-width-for-knitr-plots
#'
#' @param pl a plot object
#' @param fig.height figure height
#' @param fig.width figure width
#' @param chunk_name name of the chunk
#'
#' @author Andreas Scharmueller \email{andschar@@protonmail.com}
#'
subchunkify = function(pl,
                       fig.height = 7,
                       fig.width = 5,
                       chunk_name = 'plot') {
  pl_deparsed = paste0(deparse(function() {
    pl
  }), collapse = '')
  
  sub_chunk = paste0(
    "```{r ",
    chunk_name,
    ", fig.height=",
    fig.height,
    ", fig.width=",
    fig.width,
    ", dpi=72",
    ", echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}",
    "\n(",
    pl_deparsed,
    ")()",
    "\n```"
  )
  
  cat(knitr::knit(
    text = knitr::knit_expand(text = sub_chunk),
    quiet = TRUE
  ))
}
```

```{r, cache=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(glue)
library(Seurat)
library(scales)
library(ggh4x)
library(gridExtra)
library(pheatmap)
library(devtools)

library(CSCORE)
library(WGCNA)

ggplot2::theme_set(theme_light(base_size = 11))
opts_chunk[["set"]](
    cache = FALSE,
    dev = c("png", "pdf"),
    error = TRUE,
    highlight = TRUE,
    message = FALSE,
    prompt = FALSE,
    tidy = FALSE,
    warning = FALSE)

axis <- ggh4x::guide_axis_truncated(
  trunc_lower = unit(0, "npc"),
  trunc_upper = unit(2, "cm")
)
```

# Project details

``` {r}
set.seed(2023)

name_project <- "Co-expression of genes"
name_pi <- "Merck"

name_file <- gsub(" ", "_", name_project)
name_file <- gsub("-", "_", name_file)
```


<b>Dataset background</b>

I have pulled the dataset from the single-cell RNA-seq tutorial that we host at the core. More information about this dataset and how it processed can be found in the tutorial we have available on our website:

https://hbctraining.github.io/scRNA-seq_online/schedule/

The basic processing information for this sample includes:

- The libraries were prepared using 10X Genomics version 2 chemistry
- The samples were sequenced on the Illumina NextSeq 500
- PBMC samples from eight individual lupus patients were separated into two aliquots each.
- One aliquot of PBMCs was activated by 100 U/mL of recombinant IFN-β for 6 hours.
- The second aliquot was left untreated.
- After 6 hours, the eight samples for each condition were pooled together in two final pools (stimulated cells and control cells). We will be working with these two, pooled samples. 
- 12,138 and 12,167 cells were identified (after removing doublets) for control and stimulated pooled samples, respectively.

A brief overview of what the dataset looks like:

```{r}
# Name of column we are doing comparisons for
fill <- "sample"

n_samples <- 2
nrows <- 1
ncols <- 2
```

```{r}
# Load dataset
load(bzfile("../data/seurat_integrated.RData.bz2"))
seurat <- seurat_integrated
```

## UMAP {.tabset}

### Sample
```{r fig.width=6, fig.height=5}
Idents(seurat) <- "sample"

DimPlot(seurat, reduction = "umap") +
    xlab("UMAP1") +
    ylab("UMAP2") +
    ggtitle(glue("{name_project}\nSample")) +
    theme(plot.title = element_text(hjust = 0.5)) +
    theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(),
        axis.text.y=element_blank(), axis.ticks.y=element_blank(),
        axis.title=element_text(size=8, face="bold")) +
    guides(x = axis, y = axis) +
    theme(axis.line = element_line(arrow = arrow(type = "closed"), size = 1),
        axis.title = element_text(hjust = 0))
```


### Celltype
```{r fig.width=7, fig.height=5}
# Rename all identities
Idents(object = seurat) <- "integrated_snn_res.0.8"
seurat <- RenameIdents(object = seurat, 
                               "0" = "Naive or memory CD4+ T cells",
                               "1" = "CD14+ monocytes",
                               "2" = "Activated T cells",
                               "3" = "CD14+ monocytes",
                               "4" = "Stressed cells / Unknown",
                               "5" = "CD8+ T cells",
                               "6" = "Naive or memory CD4+ T cells",
                               "7" = "B cells",
                               "8" = "NK cells",
                               "9" = "CD8+ T cells",
                               "10" = "FCGR3A+ monocytes",
                               "11" = "B cells",
                               "12" = "NK cells",
                               "13" = "B cells",
                               "14" = "Conventional dendritic cells",
                               "15" = "Megakaryocytes",
			       "16" = "Plasmacytoid dendritic cells")
seurat$celltype <- Idents(seurat)
```

```{r}
Idents(seurat) <- "celltype"
p <- DimPlot(seurat, reduction = "umap") +
    xlab("UMAP1") +
    ylab("UMAP2") +
    ggtitle(glue("{name_project}\nSeurat Clusters")) +
    theme(plot.title = element_text(hjust = 0.5)) +
    theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(),
        axis.text.y=element_blank(), axis.ticks.y=element_blank(),
        axis.title=element_text(size=8, face="bold")) +
    guides(x = axis, y = axis) +
    theme(axis.line = element_line(arrow = arrow(type = "closed"), size = 1),
        axis.title = element_text(hjust = 0))
LabelClusters(p, id = "ident",  fontface = "bold", size = 3, bg.colour = "white", bg.r = .2, force = 0)
```

# Surfaceome genes
```{r}
surfaceome <- read.csv("../data/surfaceome.csv")

library('biomaRt')
mart <- useDataset("hsapiens_gene_ensembl", useMart("ensembl"))
gene_IDs <- getBM(filters= "ensembl_gene_id", attributes= c("ensembl_gene_id","hgnc_symbol"),
              values = surfaceome$ENSEMBL.Gene.ID, mart= mart)

surfaceome_genes <- left_join(x = surfaceome, y = gene_IDs, by = c("ENSEMBL.Gene.ID" = "ensembl_gene_id"))
surfaceome_genes <- surfaceome_genes$hgnc_symbol

surfaceome_genes <- rownames(seurat$RNA)[rownames(seurat$RNA) %in% surfaceome_genes]
as.data.frame(surfaceome_genes)
```

# Installing CS-CORE

The installation process for the tool is very straightforward and I had no issues getting it working. 

```{r}
# ## Load devtools for installing R packages from Github
# library(devtools)
# ## Install CS-CORE from Github
# install_github("ChangSuBiostats/CS-CORE")
```

# Running CS-CORE

In the vignette for [CS-CORE](https://changsubiostats.github.io/CS-CORE/articles/CSCORE.html) they subset their dataset by a singular celltype and then input their correlation matrix into [WGCNA](https://cran.r-project.org/web/packages/WGCNA/index.html) in order to determine modules of genes. 

```{r fig.height=8}
celltypes <- levels(seurat$celltype)
celltypes <- celltypes[celltypes != "Stressed cells / Unknown"]

Idents(seurat) <- "celltype"
seurat@meta.data %>% 
    ggplot() +
    geom_bar(aes(
        x = Idents(seurat),
        fill = Idents(seurat)), stat = "count", color = "black", show.legend = FALSE) +
    theme_classic() +
    xlab("Celltype") +
    ylab("Number of Cells") +
    ggtitle(glue("{name_project}\nManual celltypes")) +
    theme(plot.title = element_text(hjust = 0.5)) +
    geom_text(aes(x = Idents(seurat), label = after_stat(count)), stat = "count", vjust = -0.5) +
    theme(axis.text.x = element_text(angle = 45, hjust=1))
```

# Co-expression by celltypes

```{r results="asis", echo=FALSE}
for (ct in celltypes) {

    cat(paste0("## ", ct, "\n\n"))
    cat(paste0("To save computation time, I remove genes that have 0 counts in the ", ct, " subsetted dataset.\n"))
    seurat_sub <- subset(seurat, subset=(celltype == ct))

    # Removing genes that have 0 counts across all cells of the celltype
    counts <- seurat_sub[["RNA"]]@counts
    genes.use <- rowSums(counts) > 0
    genes.use <- names(genes.use[genes.use])
    seurat_sub <- seurat_sub[genes.use, ]

    # Surfaceome genes in the dataset
    genes <- surfaceome_genes[surfaceome_genes %in% row.names(seurat_sub)]
    n_genes <- length(genes)

    cat(paste0("After this gene filtration, we have ", n_genes, " genes from the surfaceome to use as input for CS-CORE.\n\n"))

    # Run CS-CORE
    CSCORE_result <- CSCORE(seurat_sub, genes = genes)



    cat("\n\n### Co-expression heatmap\n\n")
    cat(paste0("Here we can see the estimated correlation estimates within the ", ct, " cells.\n\n"))
    p <- pheatmap(CSCORE_result$est, show_colnames = FALSE, show_rownames = FALSE, silent=TRUE)
    subchunkify(p, n_genes/10, n_genes/10, glue("{ct}-coexp"))
    cat("\n\n")



    cat("### BH-adjusted matrix\n\n")
    cat("The vignette then has us replace gene pairs that are no significant to be 0 within the estimated co-expression matrix. This step is necessary in order to feed the CS-CORE co-expression estimated into WGCNA to determine gene modules as it originally takes in a pearson correlation matrix.\n\n") 
    cat("In essence, we are setting the co-expression values of non-significant gene pairs to be 0 - which in the context of a pearson correlation matrix would mean that the pairs are not correlated and therefore will not have co-expression.\n\n")
    cat("The modified matrix now looks like this, with the 0'd values:\n\n")

    # Obtain CS-CORE co-expression estimates
    CSCORE_coexp <- CSCORE_result$est

    # Obtain BH-adjusted p values
    CSCORE_p <- CSCORE_result$p_value
    p_matrix_BH <- matrix(0, length(genes), length(genes))
    p_matrix_BH[upper.tri(p_matrix_BH)] <- p.adjust(CSCORE_p[upper.tri(CSCORE_p)], method = "BH")
    p_matrix_BH <- p_matrix_BH + t(p_matrix_BH)

    # Set co-expression entires with BH-adjusted p-values greater than 0.05 to 0
    CSCORE_coexp[p_matrix_BH > 0.05] <- 0

    p <- pheatmap(CSCORE_coexp, show_colnames = FALSE, show_rownames = FALSE, silent=TRUE)
    subchunkify(p, n_genes/10, n_genes/10, glue("{ct}-coexp_zero"))
    cat("\n\n")



    cat("### WGCNA\n\n")
    cat("We then feed this 0'd co-expression matrix into WGCNA.\n\n")
    # Compute the adjacency matrix based on the co-expression matrix
    adj <- WGCNA::adjacency.fromSimilarity(abs(CSCORE_coexp), power = 1)
    # Compute the topological overlap matrix
    TOM <- WGCNA::TOMsimilarity(adj)
    dissTOM <- 1-TOM
    rownames(dissTOM) <- colnames(dissTOM) <- genes
    # Run hierarchical clustering as in the WGCNA workflow
    hclust_dist <- hclust(as.dist(dissTOM), method = "average") 
    memb <- dynamicTreeCut::cutreeDynamic(dendro = hclust_dist, 
                        distM = dissTOM, 
                        deepSplit = 2,
                        pamRespectsDendro = FALSE,
                        minClusterSize = 10)
    # For more instructions on how to tune the parameters in the WGCNA workflow,
    # please refer to https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/Rpackages/WGCNA/Tutorials/

    names(memb) <- genes
    memb_tab <- table(memb)
    module_list <- lapply(sort(unique(memb)), function(i_k) names(which(memb == i_k)))

    module_df <- data.frame(memb)
    module_df$memb <- as.character(module_df$memb)
    df <- module_df
    df <- tibble::rownames_to_column(df, "gene")
    colnames(df) <- c("gene", "module")

    cat(knitr::knit_print(df))
    cat("\n\n\n")
    cat("We are now given a list of modules with groups of genes that are co-expressed together in our dataset.\n\n")



    cat("### CS-CORE correlation estimates for modules\n")
    cat("Please note that this is the same correlation matrix in the section 'Co-expression heatmap', but now with annotations for the modules those genes belong to. There are no overlaps of genes among the modules.\n\n")

    module_df$genes <- row.names(module_df)
    anno <- module_df[, c("memb"), drop=F]
    # Co-expression matrix of only genes in the module
    coex_est <- CSCORE_result$est[module_df$genes, module_df$genes]
    p <- pheatmap(coex_est, show_colnames=F, 
            show_rownames=F, annotation_col=anno, annotation_row=anno, silent=T)
    subchunkify(p, n_genes/10, n_genes/10, glue("{ct}-module_coex"))
    cat("\n\n")

    # Get coexpression pairs estimate
    coex_df <- reshape2::melt(coex_est)
    colnames(coex_df) <- c("Var1", "Var2", "coex_est")

    # Get p-value of coexpression pairs
    p_est <- CSCORE_result$p_value[module_df$genes, module_df$genes]
    p_est <- reshape2::melt(p_est)
    coex_df$p_value <- p_est$value

    # Remove pairs with coexpression = 0
    # Remove pairs of the same gene (correlation will automatically be 1)
    coex_df <- coex_df %>% 
            filter(coex_est != 0) %>% 
            filter(Var1 != Var2)

    # Add module information
    coex_df$Var1_module <- coex_df$Var1
    coex_df$Var2_module <- coex_df$Var2
    for (idx in 1:nrow(module_df)) {
        module <- module_df[idx, ]$memb
        gene <- module_df[idx, ]$genes

        coex_df$Var1_module <- str_replace(coex_df$Var1_module, paste0(gene, "$"), module)
        coex_df$Var2_module <- str_replace(coex_df$Var2_module, paste0(gene, "$"), module)
    }

    # Remove pairs that are the same but ordered differently
    # Ex. (Var1, Var2) = (Var2, Var1)
    pairs <- c()
    for (idx in 1:nrow(coex_df)) {
        pair <- sort(c(coex_df[idx, ]$Var1, coex_df[idx, ]$Var2))
        pair <- paste0(pair[1], "+", pair[2])
        pairs <- c(pairs, c(pair))
    }
    coex_df$pairs <- pairs
    coex_df <- coex_df[!duplicated(coex_df$pairs), ]

    # Save output
    write.csv(coex_df, glue("results-cs_core/{ct}.csv"))



    cat("### Top co-expressed pairs\n\n")
    cat(glue("I have removed gene pairs that are not significant (0.05) and sorted the values by p-value to more easily identify interesting gene pairs within {ct}. I have also written out the results as a csv file before this filtration in case there are other thresholds we want to explore.\n\n"))
    # Remove non-significant correlations
    coex_df <- coex_df %>%
        filter(p_value <= 0.05) %>%
        arrange(p_value)
    cat(knitr::knit_print(coex_df))
    cat("\n\n\n")



    cat("### Scatterplot top pair\n\n")
    cat(glue("As a sanity check, I wanted to visualize the expression of the top correlated pair within the {ct} population.\n\n"))

    # Gene name of top co-expressed genes
    gene_1 <- as.character(coex_df[1, "Var1"])
    gene_2 <- as.character(coex_df[1, "Var2"])
    # Counts for genes
    counts <- data.frame(t(data.frame(seurat_sub$RNA@data[c(gene_1, gene_2), ])))
    colnames(counts) <- c(gene_1, gene_2)

    # Scatterplot
    scatter <- ggplot(counts) +
        geom_point(aes(x=get(gene_1), y=get(gene_2))) +
        scale_fill_continuous(type = "viridis") +
        theme_classic() +
        xlab(gene_1) +
        ylab(gene_2) +
        geom_abline(slope=1, intercept = 0, color="red")

    n_zero <- dim(counts)[1]
    # Remove cells with 0 expression for both genes
    counts_zero <- counts[rowSums(counts) > 0, ]
    n_zero <- n_zero - dim(counts_zero)[1]

    cat(glue("As the scatterplot can make it difficult to see how many cells overlap at the same values, I have taken the same information displayed as a density plot. To create a better scale, I removed cells that have 0 expression for both genes ({n_zero} cells).\n\n"))

    # Density scatterplot
    density <- ggplot(counts_zero) +
        geom_bin2d(aes(x=get(gene_1), y=get(gene_2)), bins=70) +
        scale_fill_continuous(type = "viridis") +
        theme_classic() +
        xlab(gene_1) +
        ylab(gene_2) +
        geom_abline(slope=1, intercept = 0, color="red")

    p <- cowplot::plot_grid(scatter, density, ncol=2)
    subchunkify(p, 6, 13, glue("{ct}-scatter"))
    cat("\n\n")
}
```


# Notes

CS-CORE seems to be assigning correlation values of ~1 to gene pairs that have high expression of each gene - regardless of whether or not they are co-expressed. This can be best seen in the scatterplots where we see 0 expression for 1 gene pair but high expression for the other and vice versa. This could potentially be a case of droput where we are not able to always capture expression from both genes, but when they do they are highly correlated. If this is the case, ideally imputation would be able to counteract this effect. Alternatively I also think we can filter correlation pairs based on percentages of cells that express both genes within the population that we are looking at.

Additionally, the results from WGCNA are not necessarily concordant with the coexpression patterns, as we can see in the heatmaps that genes of the same module do not necessarily cluster cleanly together.

Shannan had the suggestion of using a Multiple T test correction to account for the "significant" correlated pairs that do not seem to be correct based on the gene expression (ex. CAVIN2 and ENO1 in the Plasmacytoid dendritic cells).

# Next steps

Benchmarking these results will be the next focal point of the analysis. James has offered a dataset that he thinks would be good to use as a ground truth. 

I am going to focus on looking at the clustering accuracy within gene modules, using metrics such as: adjusted rand index, subspace distance between PCs, spearman correlation. These are all metrics that the CS-CORE paper used for benchmarking.