---
title: "Merck - CS-CORE PBMC 3k"
author: "Noor Sohail"
output:
   html_document:
      code_folding: hide
      df_print: paged
      highlights: pygments
      number_sections: true
      self_contained: true
      theme: default
      toc: true
      toc_float:
         collapsed: true
         smooth_scroll: true
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
# Turn off Warnings and other console output messages from the whole document
```

```{r}
#' Create sub-chunks for plots
#' Plotting function to generate plots of a specific dimension when html is knitted
#'
#' taken from: https://stackoverflow.com/questions/15365829/dynamic-height-and-width-for-knitr-plots
#'
#' @param pl a plot object
#' @param fig.height figure height
#' @param fig.width figure width
#' @param chunk_name name of the chunk
#'
#' @author Andreas Scharmueller \email{andschar@@protonmail.com}
#'
subchunkify = function(pl,
                       fig.height = 7,
                       fig.width = 5,
                       chunk_name = 'plot') {
  pl_deparsed = paste0(deparse(function() {
    pl
  }), collapse = '')
  
  sub_chunk = paste0(
    "```{r ",
    chunk_name,
    ", fig.height=",
    fig.height,
    ", fig.width=",
    fig.width,
    ", dpi=72",
    ", echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}",
    "\n(",
    pl_deparsed,
    ")()",
    "\n```"
  )
  
  cat(knitr::knit(
    text = knitr::knit_expand(text = sub_chunk),
    quiet = TRUE
  ))
}
```

```{r, cache=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(glue)
library(Seurat)
library(scales)
library(ggh4x)
library(gridExtra)
library(pheatmap)
library(devtools)

library(CSCORE)
library(WGCNA)

ggplot2::theme_set(theme_light(base_size = 11))
opts_chunk[["set"]](
    cache = FALSE,
    dev = c("png", "pdf"),
    error = TRUE,
    highlight = TRUE,
    message = FALSE,
    prompt = FALSE,
    tidy = FALSE,
    warning = FALSE)

axis <- ggh4x::guide_axis_truncated(
  trunc_lower = unit(0, "npc"),
  trunc_upper = unit(2, "cm")
)
```

# Project details

``` {r}
set.seed(2023)

name_project <- "Co-expression of genes"
name_pi <- "Merck"

name_file <- gsub(" ", "_", name_project)
name_file <- gsub("-", "_", name_file)
```


<b>Dataset background</b>

I have pulled the dataset from the single-cell RNA-seq tutorial that we host at the core. More information about this dataset and how it processed can be found in the tutorial we have available on our website:

https://hbctraining.github.io/scRNA-seq_online/schedule/

The basic processing information for this sample includes:

- The libraries were prepared using 10X Genomics version 2 chemistry
- The samples were sequenced on the Illumina NextSeq 500
- PBMC samples from eight individual lupus patients were separated into two aliquots each.
- One aliquot of PBMCs was activated by 100 U/mL of recombinant IFN-β for 6 hours.
- The second aliquot was left untreated.
- After 6 hours, the eight samples for each condition were pooled together in two final pools (stimulated cells and control cells). We will be working with these two, pooled samples. 
- 12,138 and 12,167 cells were identified (after removing doublets) for control and stimulated pooled samples, respectively.

A brief overview of what the dataset looks like:

```{r}
# Name of column we are doing comparisons for
fill <- "sample"

n_samples <- 2
nrows <- 1
ncols <- 2
```

```{r}
# Load dataset
load(bzfile("../data/seurat_integrated.RData.bz2"))
seurat <- seurat_integrated
```

## UMAP {.tabset}

### Sample
```{r fig.width=6, fig.height=5}
Idents(seurat) <- "sample"

DimPlot(seurat, reduction = "umap") +
    xlab("UMAP1") +
    ylab("UMAP2") +
    ggtitle(glue("{name_project}\nSample")) +
    theme(plot.title = element_text(hjust = 0.5)) +
    theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(),
        axis.text.y=element_blank(), axis.ticks.y=element_blank(),
        axis.title=element_text(size=8, face="bold")) +
    guides(x = axis, y = axis) +
    theme(axis.line = element_line(arrow = arrow(type = "closed"), size = 1),
        axis.title = element_text(hjust = 0))
```


### Celltype
```{r fig.width=7, fig.height=5}
# Rename all identities
Idents(object = seurat) <- "integrated_snn_res.0.8"
seurat <- RenameIdents(object = seurat, 
                               "0" = "Naive or memory CD4+ T cells",
                               "1" = "CD14+ monocytes",
                               "2" = "Activated T cells",
                               "3" = "CD14+ monocytes",
                               "4" = "Stressed cells / Unknown",
                               "5" = "CD8+ T cells",
                               "6" = "Naive or memory CD4+ T cells",
                               "7" = "B cells",
                               "8" = "NK cells",
                               "9" = "CD8+ T cells",
                               "10" = "FCGR3A+ monocytes",
                               "11" = "B cells",
                               "12" = "NK cells",
                               "13" = "B cells",
                               "14" = "Conventional dendritic cells",
                               "15" = "Megakaryocytes",
			       "16" = "Plasmacytoid dendritic cells")
seurat$celltype <- Idents(seurat)
```

```{r}
Idents(seurat) <- "celltype"
p <- DimPlot(seurat, reduction = "umap") +
    xlab("UMAP1") +
    ylab("UMAP2") +
    ggtitle(glue("{name_project}\nSeurat Clusters")) +
    theme(plot.title = element_text(hjust = 0.5)) +
    theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(),
        axis.text.y=element_blank(), axis.ticks.y=element_blank(),
        axis.title=element_text(size=8, face="bold")) +
    guides(x = axis, y = axis) +
    theme(axis.line = element_line(arrow = arrow(type = "closed"), size = 1),
        axis.title = element_text(hjust = 0))
LabelClusters(p, id = "ident",  fontface = "bold", size = 3, bg.colour = "white", bg.r = .2, force = 0)
```

# Surfaceome genes
```{r}
surfaceome <- read.csv("../data/surfaceome.csv")

library('biomaRt')
mart <- useDataset("hsapiens_gene_ensembl", useMart("ensembl"))
gene_IDs <- getBM(filters= "ensembl_gene_id", attributes= c("ensembl_gene_id","hgnc_symbol"),
              values = surfaceome$ENSEMBL.Gene.ID, mart= mart)

surfaceome_genes <- left_join(x = surfaceome, y = gene_IDs, by = c("ENSEMBL.Gene.ID" = "ensembl_gene_id"))
surfaceome_genes <- surfaceome_genes$hgnc_symbol

surfaceome_genes <- rownames(seurat$RNA)[rownames(seurat$RNA) %in% surfaceome_genes]
as.data.frame(surfaceome_genes)
```

# Installing CS-CORE

The installation process for the tool is very straightforward adn I had no issues getting it working. 

```{r}
# ## Load devtools for installing R packages from Github
# library(devtools)
# ## Install CS-CORE from Github
# install_github("ChangSuBiostats/CS-CORE")
```

# Running CS-CORE

In the vignette for CS-CORE they subset their dataset by a singular celltype. 


```{r fig.height=8}
celltypes <- levels(seurat$celltype)
celltypes <- celltypes[celltypes != "Stressed cells / Unknown"]

Idents(seurat) <- "celltype"
seurat@meta.data %>% 
    ggplot() +
    geom_bar(aes(
        x = Idents(seurat),
        fill = Idents(seurat)), stat = "count", color = "black", show.legend = FALSE) +
    theme_classic() +
    xlab("Celltype") +
    ylab("Number of Cells") +
    ggtitle(glue("{name_project}\nManual celltypes")) +
    theme(plot.title = element_text(hjust = 0.5)) +
    geom_text(aes(x = Idents(seurat), label = after_stat(count)), stat = "count", vjust = -0.5) +
    theme(axis.text.x = element_text(angle = 45, hjust=1))
```

# Co-expression by celltypes

```{r results="asis", echo=FALSE}
for (ct in celltypes) {

  cat(paste0("## ", ct, "\n\n"))
  cat(paste0("To save computationtime, I remove genes that have 0 counts in the ", ct, " subsetted dataset.\n"))
  seurat_sub <- subset(seurat, subset=(celltype == ct))

  # Removing genes that have 0 counts across all cells of the celltype
  counts <- seurat_sub[["RNA"]]@counts
  genes.use <- rowSums(counts) > 0
  genes.use <- names(genes.use[genes.use])
  seurat_sub <- seurat_sub[genes.use, ]

  # Surfaceome genes in the dataset
  genes <- surfaceome_genes[surfaceome_genes %in% row.names(seurat_sub)]
  n_genes <- length(genes)

  cat(paste0("After this gene filtration, we have ", n_genes, " genes from the surfaceome to use as input for CS-CORE.\n\n"))

  # Run CS-CORE
  CSCORE_result <- CSCORE(seurat_sub, genes = genes)

  cat("\n\n### Coexpression heatmap\n\n")
  cat(paste0("Here we can see the estimated correlation scores for the full set of genes within the ", ct, " cells.\n\n"))
  p <- pheatmap(CSCORE_result$est, show_colnames = FALSE, show_rownames = FALSE, silent=TRUE)
  subchunkify(p, n_genes/10, n_genes/10, glue("{ct}-coexp"))
  cat("\n\n")

  cat("### BH-adjusted matrix\n\n")

  cat("The vignette then has us replace gene pairs that are no significant to be 0 within the estimated co-expression matrix. This step is necessary in order to feed the CS-CORE co-expression estimated into WGCNA to determine gene modules as it originally takes in a pearson correlation matrix.\n\n") 
  cat("In essence, we are setting the co-expression values of non-significant gene pairs to be 0 - which in the context of a pearson correlation matrix would mean that the pairs are not correlated and therefore will not have co-expression.\n\n")
  cat("The modified matrix now looks like this, with the 0'd values:\n\n")
  
  # Obtain CS-CORE co-expression estimates
  CSCORE_coexp <- CSCORE_result$est

  # Obtain BH-adjusted p values
  CSCORE_p <- CSCORE_result$p_value
  p_matrix_BH <- matrix(0, length(genes), length(genes))
  p_matrix_BH[upper.tri(p_matrix_BH)] <- p.adjust(CSCORE_p[upper.tri(CSCORE_p)], method = "BH")
  p_matrix_BH <- p_matrix_BH + t(p_matrix_BH)

  # Set co-expression entires with BH-adjusted p-values greater than 0.05 to 0
  CSCORE_coexp[p_matrix_BH > 0.05] <- 0

  p <- pheatmap(CSCORE_coexp, show_colnames = FALSE, show_rownames = FALSE, silent=TRUE)
  subchunkify(p, n_genes/10, n_genes/10, glue("{ct}-coexp_zero"))
  cat("\n\n")

  cat("### WGCNA\n\n")
  cat("To determine gene modules and created a weighted correlation network, we feed this coexpression matrix into WGCNA. Traditionally, a pearson correlation matrix would be used as the input for WGNCA, but instead we are providing a coexpression matrix (the output from CS-CORE).\n\n")
  # Compute the adjacency matrix based on the co-expression matrix
  adj <- WGCNA::adjacency.fromSimilarity(abs(CSCORE_coexp), power = 1)
  # Compute the topological overlap matrix
  TOM <- WGCNA::TOMsimilarity(adj)
  dissTOM <- 1-TOM
  rownames(dissTOM) <- colnames(dissTOM) <- genes
  # Run hierarchical clustering as in the WGCNA workflow
  hclust_dist <- hclust(as.dist(dissTOM), method = "average") 
  memb <- dynamicTreeCut::cutreeDynamic(dendro = hclust_dist, 
                      distM = dissTOM, 
                      deepSplit = 2,
                      pamRespectsDendro = FALSE,
                      minClusterSize = 10)
  # For more instructions on how to tune the parameters in the WGCNA workflow,
  # please refer to https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/Rpackages/WGCNA/Tutorials/

  names(memb) <- genes
  memb_tab <- table(memb)
  module_list <- lapply(sort(unique(memb)), function(i_k) names(which(memb == i_k)))
  cat("\n\nWe are now given a list of modules with groups of genes that are co-expressed together in our dataset.\n\n")

  cat("### Modules\n\n")
  for (idx_module in 1:length(module_list)) {
    cat(paste0("### Module ", idx_module, "\n\n"))

    # Genes within the module
    cat("##### Genes in the module\n\n")
    module <- module_list[[idx_module]]
    n_genes <- length(module)
    cat(glue("In module {idx_module} there are `{n_genes} genes.\n\n"))

    if (n_genes < 25) {
      n_genes <- 25
    }

    cat(knitr::knit_print(data.frame(module)))
    cat("\n\n")

    cat("##### Co-expression of module genes\n\n")
    cat("Now we can take a look at the CS-CORE co-expression estimations of just the genes in this module.\n\n")

    # Co-expression matrix of only genes in the module
    coex_est <- CSCORE_result$est[module, module]
    p <- pheatmap(coex_est, show_colnames = F, show_rownames = F, silent=T)
    subchunkify(p, n_genes/5, n_genes/5, glue("{ct}-{idx_module}-heatmap"))
    cat("\n\n")

    # Get coexpression pairs
    coex_df <- reshape2::melt(coex_est)
    coex_df <- coex_df %>% 
                filter(value != 0) %>% 
                filter(Var1 != Var2) %>%
                arrange(desc(value))

    # Co-expression values are all 0
    if(dim(coex_df)[1] == 0) {
      next
    }

    cat("##### Top co-expressed pairs\n\n")
    cat(glue("I have sorted the co-expression estimates from the matrix as a dataframe to more easily view which genes pairs at highly co-expressed in {ct}.\n\n"))
    cat(knitr::knit_print(coex_df))
    cat("\n\n\n")

    # Gene name of top co-expressed genes
    gene_1 <- as.character(coex_df[1, "Var1"])
    gene_2 <- as.character(coex_df[1, "Var2"])
    # Counts for genes
    counts <- data.frame(t(data.frame(seurat_sub$RNA@data[c(gene_1, gene_2), ])))
    colnames(counts) <- c(gene_1, gene_2) # sometimes converts "-" to "."

    cat("##### Scatterplot top pair\n\n")
    cat(glue("As a sanity check, I wanted to visualize the expression of the top correlated pair within the {ct} population.\n\n"))
    # Scatterplot
    p <- ggplot(counts) +
        geom_point(aes(x=get(gene_1), y=get(gene_2))) +
        scale_fill_continuous(type = "viridis") +
        theme_classic() +
        xlab(gene_1) +
        ylab(gene_2) +
        geom_abline(slope=1, intercept = 0, color="red")
    subchunkify(p, 6, 6, glue("{ct}-{idx_module}-scatter"))
    cat("\n\n")

    cat("##### Density plot top pair\n\n")
    n_zero <- dim(counts)[1]
    # Remove cells with 0 expression for both genes
    counts <- counts[rowSums(counts) > 0, ]
    n_zero <- n_zero - dim(counts)[1]

    cat(glue("As the scatterplot can make it difficult to see how many cells overlap at the same values, I have taken the same information displayed as a density plot. To create a better scale, I removed cells that have 0 expression for both genes ({n_zero} cells).\n\n"))

    # Density plot
    p <- ggplot(counts) +
        geom_bin2d(aes(x=get(gene_1), y=get(gene_2)), bins=70) +
        scale_fill_continuous(type = "viridis") +
        theme_classic() +
        xlab(gene_1) +
        ylab(gene_2) +
        geom_abline(slope=1, intercept = 0, color="red")
    subchunkify(p, 6, 6, glue("{ct}-{idx_module}-density"))
    cat("\n\n")
  }
}
```

# Note

CS-CORE seems to be assigning correlation values of ~1 to gene pairs that have high expression of each gene - regardless of whether or not they are co-expressed. This can be best seen in the scatterplots where we see 0 expression for 1 gene pair but high expression for the other and vice versa.

# Next steps

Benchmarking of CS-CORE will be the next steps. I am going to focus on looking at the clustering accuracy within the gene modules generated by WGCNA, using metrics such as: adjusted rand index, subspace distance between PCs, spearman correlation).

Additionally, this would be a good time to start using a ground truth dataset - such as the FISH or CITE-seq datasets we were discussing.