---
title: "Merck - MAGIC PBMC 3k"
author: "Noor Sohail"
output:
   html_document:
      code_folding: hide
      df_print: paged
      highlights: pygments
      number_sections: true
      self_contained: true
      theme: default
      toc: true
      toc_float:
         collapsed: true
         smooth_scroll: true
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
# Turn off Warnings and other console output messages from the whole document
```

```{r}
#' Create sub-chunks for plots
#' Plotting function to generate plots of a specific dimension when html is knitted
#'
#' taken from: https://stackoverflow.com/questions/15365829/dynamic-height-and-width-for-knitr-plots
#'
#' @param pl a plot object
#' @param fig.height figure height
#' @param fig.width figure width
#' @param chunk_name name of the chunk
#'
#' @author Andreas Scharmueller \email{andschar@@protonmail.com}
#'
subchunkify = function(pl,
                       fig.height = 7,
                       fig.width = 5,
                       chunk_name = 'plot') {
  pl_deparsed = paste0(deparse(function() {
    pl
  }), collapse = '')
  
  sub_chunk = paste0(
    "```{r ",
    chunk_name,
    ", fig.height=",
    fig.height,
    ", fig.width=",
    fig.width,
    ", dpi=72",
    ", echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}",
    "\n(",
    pl_deparsed,
    ")()",
    "\n```"
  )
  
  cat(knitr::knit(
    text = knitr::knit_expand(text = sub_chunk),
    quiet = TRUE
  ))
}
```

```{r, cache=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(glue)
library(Seurat)
library(scales)
library(ggh4x)
library(gridExtra)
library(pheatmap)
library(devtools)

ggplot2::theme_set(theme_light(base_size = 11))
opts_chunk[["set"]](
    cache = FALSE,
    dev = c("png", "pdf"),
    error = TRUE,
    highlight = TRUE,
    message = FALSE,
    prompt = FALSE,
    tidy = FALSE,
    warning = FALSE)

axis <- ggh4x::guide_axis_truncated(
  trunc_lower = unit(0, "npc"),
  trunc_upper = unit(2, "cm")
)
```

# Project details

``` {r}
set.seed(2023)

name_project <- "Co-expression of genes"
name_pi <- "Merck"

name_file <- gsub(" ", "_", name_project)
name_file <- gsub("-", "_", name_file)
```


<b>Dataset background</b>

I have pulled the dataset from the single-cell RNA-seq tutorial that we host at the core. More information about this dataset and how it processed can be found in the tutorial we have available on our website:

https://hbctraining.github.io/scRNA-seq_online/schedule/

The basic processing information for this sample includes:

- The libraries were prepared using 10X Genomics version 2 chemistry
- The samples were sequenced on the Illumina NextSeq 500
- PBMC samples from eight individual lupus patients were separated into two aliquots each.
- One aliquot of PBMCs was activated by 100 U/mL of recombinant IFN-β for 6 hours.
- The second aliquot was left untreated.
- After 6 hours, the eight samples for each condition were pooled together in two final pools (stimulated cells and control cells). We will be working with these two, pooled samples. 
- 12,138 and 12,167 cells were identified (after removing doublets) for control and stimulated pooled samples, respectively.

A brief overview of what the dataset looks like:

```{r}
# Name of column we are doing comparisons for
fill <- "sample"

n_samples <- 2
nrows <- 1
ncols <- 2
```

```{r}
# Load dataset
load(bzfile("../data/seurat_integrated.RData.bz2"))
seurat <- seurat_integrated
```

## UMAP {.tabset}

### Sample
```{r fig.width=6, fig.height=5}
Idents(seurat) <- "sample"

DimPlot(seurat, reduction = "umap") +
    xlab("UMAP1") +
    ylab("UMAP2") +
    ggtitle(glue("{name_project}\nSample")) +
    theme(plot.title = element_text(hjust = 0.5)) +
    theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(),
        axis.text.y=element_blank(), axis.ticks.y=element_blank(),
        axis.title=element_text(size=8, face="bold")) +
    guides(x = axis, y = axis) +
    theme(axis.line = element_line(arrow = arrow(type = "closed"), size = 1),
        axis.title = element_text(hjust = 0))
```


### Celltype
```{r fig.width=7, fig.height=5}
# Rename all identities
Idents(object = seurat) <- "integrated_snn_res.0.8"
seurat <- RenameIdents(object = seurat, 
                               "0" = "Naive or memory CD4+ T cells",
                               "1" = "CD14+ monocytes",
                               "2" = "Activated T cells",
                               "3" = "CD14+ monocytes",
                               "4" = "Stressed cells / Unknown",
                               "5" = "CD8+ T cells",
                               "6" = "Naive or memory CD4+ T cells",
                               "7" = "B cells",
                               "8" = "NK cells",
                               "9" = "CD8+ T cells",
                               "10" = "FCGR3A+ monocytes",
                               "11" = "B cells",
                               "12" = "NK cells",
                               "13" = "B cells",
                               "14" = "Conventional dendritic cells",
                               "15" = "Megakaryocytes",
			       "16" = "Plasmacytoid dendritic cells")
seurat$celltype <- Idents(seurat)

celltypes <- levels(seurat$celltype)
celltypes <- celltypes[celltypes != "Stressed cells / Unknown"]
```

```{r}
Idents(seurat) <- "celltype"
p <- DimPlot(seurat, reduction = "umap") +
    xlab("UMAP1") +
    ylab("UMAP2") +
    ggtitle(glue("{name_project}\nSeurat Clusters")) +
    theme(plot.title = element_text(hjust = 0.5)) +
    theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(),
        axis.text.y=element_blank(), axis.ticks.y=element_blank(),
        axis.title=element_text(size=8, face="bold")) +
    guides(x = axis, y = axis) +
    theme(axis.line = element_line(arrow = arrow(type = "closed"), size = 1),
        axis.title = element_text(hjust = 0))
LabelClusters(p, id = "ident",  fontface = "bold", size = 3, bg.colour = "white", bg.r = .2, force = 0)
```

# MAGIC
```{r}
# git clone git://github.com/KrishnaswamyLab/MAGIC.git
# cd MAGIC/python
# python setup.py install --user
# cd ../Rmagic
# R CMD INSTALL .

### Please see the 03_magic.R script
```


# Imputation with MAGIC by celltypes

```{r results="asis", echo=FALSE}
for (ct in celltypes) {

  cat(paste0("## ", ct, "\n\n"))
  seurat_sub <- subset(seurat, subset=(celltype == ct))

  coex_df <- read.csv(glue("02.1-cs_core/{ct}.csv")) %>% arrange(p_value)
  # Gene name of top co-expressed genes
  gene_1 <- as.character(coex_df[1, "Var1"])
  gene_2 <- as.character(coex_df[1, "Var2"])

  cat("### Before imputation\n")
  # Counts for genes
  counts <- data.frame(t(data.frame(seurat_sub$RNA@data[c(gene_1, gene_2), ])))
  colnames(counts) <- c(gene_1, gene_2)

  # Scatterplot
  scatter <- ggplot(counts) +
      geom_point(aes(x=get(gene_1), y=get(gene_2))) +
      scale_fill_continuous(type = "viridis") +
      theme_classic() +
      xlab(gene_1) +
      ylab(gene_2) +
      geom_abline(slope=1, intercept = 0, color="red")

  n_zero <- dim(counts)[1]
  # Remove cells with 0 expression for both genes
  counts_zero <- counts[rowSums(counts) > 0, ]
  n_zero <- n_zero - dim(counts_zero)[1]

  cat(glue("As the scatterplot can make it difficult to see how many cells overlap at the same values, I have taken the same information displayed as a density plot. To create a better scale, I removed cells that have 0 expression for both genes ({n_zero} cells).\n\n"))

  # Density scatterplot
  density <- ggplot(counts_zero) +
      geom_bin2d(aes(x=get(gene_1), y=get(gene_2)), bins=70) +
      scale_fill_continuous(type = "viridis") +
      theme_classic() +
      xlab(gene_1) +
      ylab(gene_2) +
      geom_abline(slope=1, intercept = 0, color="red")

  p <- cowplot::plot_grid(scatter, density, ncol=2)
  subchunkify(p, 6, 13, glue("{ct}-scatter"))
  cat("\n\n")


  cat("### After magic imputation\n")
  # Counts for genes
  counts <- read.csv(glue("03_magic/{ct}.csv"))
  gene_1 <- str_replace(gene_1, "-", ".")
  gene_2 <- str_replace(gene_2, "-", ".")
  counts <- counts[c(gene_1, gene_2)]

  colnames(counts) <- c(gene_1, gene_2)

  # Scatterplot
  scatter <- ggplot(counts) +
      geom_point(aes(x=get(gene_1), y=get(gene_2))) +
      scale_fill_continuous(type = "viridis") +
      theme_classic() +
      xlab(gene_1) +
      ylab(gene_2) +
      geom_abline(slope=1, intercept = 0, color="red")

  n_zero <- dim(counts)[1]
  # Remove cells with 0 expression for both genes
  counts_zero <- counts[rowSums(counts) > 0, ]
  n_zero <- n_zero - dim(counts_zero)[1]

  cat(glue("As the scatterplot can make it difficult to see how many cells overlap at the same values, I have taken the same information displayed as a density plot. To create a better scale, I removed cells that have 0 expression for both genes ({n_zero} cells).\n\n"))

  # Density scatterplot
  density <- ggplot(counts_zero) +
      geom_bin2d(aes(x=get(gene_1), y=get(gene_2)), bins=70) +
      scale_fill_continuous(type = "viridis") +
      theme_classic() +
      xlab(gene_1) +
      ylab(gene_2) +
      geom_abline(slope=1, intercept = 0, color="red")

  p <- cowplot::plot_grid(scatter, density, ncol=2)
  subchunkify(p, 6, 13, glue("{ct}-scatter_imputed"))
  cat("\n\n")
}
```