---
title: "Workflow overview"
author: "Noor Sohail"
output:
   html_document:
      code_folding: hide
      df_print: paged
      highlights: pygments
      number_sections: true
      self_contained: true
      theme: default
      toc: true
      toc_float:
         collapsed: true
         smooth_scroll: true
---

# Parameters to set in advance

celltype of interest
filter genes based on thresholds

```{r}
ct <- "CD8+ T cells"
filter <- TRUE
```


```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
# Turn off Warnings and other console output messages from the whole document
```

```{r, cache=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(glue)
library(Seurat)
library(scales)
library(ggh4x)
library(gridExtra)
library(pheatmap)
library(devtools)

library(CSCORE)
library(WGCNA)
library(ggpubr)
library(ggvenn)

library(SAVER)

ggplot2::theme_set(theme_light(base_size = 11))
opts_chunk[["set"]](
    cache = FALSE,
    dev = c("png", "pdf"),
    error = TRUE,
    highlight = TRUE,
    message = FALSE,
    prompt = FALSE,
    tidy = FALSE,
    warning = FALSE)

theme_arrow <- function() {
    axis <- ggh4x::guide_axis_truncated(
        trunc_lower = unit(0, "npc"),
        trunc_upper = unit(3, "cm")
    )

    list(
        theme_classic() +
        theme(plot.title=element_text(hjust = 0.5, face="bold")) +
        theme(axis.text.x=element_blank(), axis.ticks.x=element_blank()) +
        theme(axis.text.y=element_blank(), axis.ticks.y=element_blank()) +
        theme(axis.title=element_text(size=10, face="bold", hjust=0)) +
        theme(axis.line = element_line(arrow = arrow(type="closed", length=unit(0.5, "cm")), linewidth = 1)),
        guides(x=axis, y=axis)
    )
}

#' Create sub-chunks for plots
#'
#' taken from: https://stackoverflow.com/questions/15365829/dynamic-height-and-width-for-knitr-plots
#'
#' @param pl a plot object
#' @param fig.height figure height
#' @param fig.width figure width
#' @param chunk_name name of the chunk
#'
#' @author Andreas Scharmueller \email{andschar@@protonmail.com}
#'
subchunkify = function(pl,
                       fig.height = 7,
                       fig.width = 5,
                       chunk_name = 'plot') {
  pl_deparsed = paste0(deparse(function() {
    pl
  }), collapse = '')
  
  sub_chunk = paste0(
    "```{r ",
    chunk_name,
    ", fig.height=",
    fig.height,
    ", fig.width=",
    fig.width,
    ", dpi=72",
    ", echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}",
    "\n(",
    pl_deparsed,
    ")()",
    "\n```"
  )
  
  cat(knitr::knit(
    text = knitr::knit_expand(text = sub_chunk),
    quiet = TRUE
  ))
  cat("\n\n")
}
```

# Overview

The overaching goal of this project is to assess different ways of identifying novel co-expressed gene pairs from single-cell RNA-seq datasets. We know that correlations based on the normalized RNA expression are not accurate due to dropout. 


# Project details

``` {r}
set.seed(2023)

name_project <- "Co-expression of genes"
name_pi <- "Merck"

name_file <- gsub(" ", "_", name_project)
name_file <- gsub("-", "_", name_file)
```

```{r}
path_outs <- "04.0_overview_figures"
```

<b>Dataset background</b>

I have pulled the dataset from the single-cell RNA-seq tutorial that we host at the core. More information about this dataset and how it processed can be found in the tutorial we have available on our website:

https://hbctraining.github.io/scRNA-seq_online/schedule/

The basic processing information for this sample includes:

- The libraries were prepared using 10X Genomics version 2 chemistry
- The samples were sequenced on the Illumina NextSeq 500
- PBMC samples from eight individual lupus patients were separated into two aliquots each.
- One aliquot of PBMCs was activated by 100 U/mL of recombinant IFN-β for 6 hours.
- The second aliquot was left untreated.
- After 6 hours, the eight samples for each condition were pooled together in two final pools (stimulated cells and control cells). We will be working with these two, pooled samples. 
- 12,138 and 12,167 cells were identified (after removing doublets) for control and stimulated pooled samples, respectively.

A brief overview of what the dataset looks like:


```{r}
# Load dataset
# remotes::install_github("mojaveazure/seurat-disk", quiet=F)
library(SeuratDisk)
load(bzfile("../data/seurat_integrated.RData.bz2"))

seurat <- seurat_integrated
DefaultAssay(seurat) <- "RNA"
seurat[["SCT"]] <- NULL
seurat[["integrated"]] <- NULL

rm(seurat_integrated)
```

```{r}
Idents(object = seurat) <- "integrated_snn_res.0.8"

seurat <- RenameIdents(object = seurat, 
                               "0" = "Naive or memory CD4+ T cells",
                               "1" = "CD14+ monocytes",
                               "2" = "Activated T cells",
                               "3" = "CD14+ monocytes",
                               "4" = "Stressed cells / Unknown",
                               "5" = "CD8+ T cells",
                               "6" = "Naive or memory CD4+ T cells",
                               "7" = "B cells",
                               "8" = "NK cells",
                               "9" = "CD8+ T cells",
                               "10" = "FCGR3A+ monocytes",
                               "11" = "B cells",
                               "12" = "NK cells",
                               "13" = "B cells",
                               "14" = "Conventional dendritic cells",
                               "15" = "Megakaryocytes",
			       "16" = "Plasmacytoid dendritic cells")

seurat$celltype <- Idents(seurat)
Idents(seurat) <- "celltype"
```

```{r fig.height=6}
seurat@meta.data %>%
        ggplot() +
        geom_bar(aes(
            x = celltype,
            fill=celltype),
            stat = "count", color = "black") +
        theme_classic() +
        NoLegend() +
        xlab("Celltype") +
        ylab("Number of Cells") +
        ggtitle(glue("{name_project}\nCelltypes")) +
        theme(plot.title = element_text(hjust = 0.5)) +
        geom_text(aes(x = celltype, label = after_stat(count)), stat = "count", vjust = -0.5) +
        theme(axis.text.x = element_text(angle = 45, hjust=1))
```

```{r fig.height=5, fig.width=7}
Idents(seurat) <- "celltype"
p <- DimPlot(seurat, reduction = "umap") +
    theme_arrow() +
    ggtitle(glue("{name_project}\nCelltypes"))
LabelClusters(p, id = "ident",  fontface = "bold", size = 3, bg.colour = "white", bg.r = .2, force = 0)
```

# Zooming in on `r ct` and surfaceome genes

## Subset to celltype

```{r}
seurat <- subset(seurat, celltype == ct)

# Removing genes that have 0 counts across all cells of the celltype
counts <- seurat[["RNA"]]@counts
genes.use <- rowSums(counts) > 0
genes.use <- names(genes.use[genes.use])
seurat <- seurat[genes.use, ]
rm(counts)
```

```{r}
# saveRDS(seurat, glue("{path_outs}/seurat.RDS"))

seurat <- readRDS(glue("{path_outs}/seurat.RDS"))
```

## Surfaceome genes

```{r}
surfaceome_genes <- read.csv("../data/surfaceome_genes.csv")$x

# Get the updated list of surfaceome genes found in the dataset
genes.use <- rownames(seurat)
surfaceome_genes_all <- surfaceome_genes[surfaceome_genes %in% genes.use]
as.data.frame(surfaceome_genes_all)
```

## Known markers

Using some known markers as positive and negative controls for validation.

```{r }
markers <- list()

markers[["CD8"]] <- c("CCL5", "CST7", "GZMB", "CD8A", "PRF1", "CD3D")
markers[["B"]] <- c("CD79A", "MS4A1", "IGHM")
markers[["Monocytes"]] <- c("CD14", "LYZ", "FCGR3A", "MS4A7")

genes <- unname(unlist(markers))

# Dataframe of gene markers
# For clearer plots later on
genes_label <- c()
for (marker in names(markers)) {
  g <- markers[[marker]]
  names(g) <- rep(marker, length(g))

  genes_label <- c(genes_label, g)
}
genes_label <- data.frame(gene=genes_label, celltype=names(genes_label)) %>% column_to_rownames(var="gene")
anno_colors <- list(celltype = c("B"="red", "CD8"="blue", "Monocytes"="green"))

genes_label
```


# The dropout problem

```{r fig.height=9, fig.width=15}
# Get RNA expression counts
data_rna <- FetchData(seurat[["RNA"]], vars=genes)

# Unique combination of each gene pair
genes_comb <- data.frame(t(combn(markers[["CD8"]], 2)))

plot_list <- list()
for (idx in 1:nrow(genes_comb)) {
    gene_1 <- genes_comb[idx, 1]
    gene_2 <- genes_comb[idx, 2]

    # Find cell IDs for cells with dropout
    dropout_id_1 <- row.names(data_rna[(data_rna[gene_1] == 0.0), ])
    dropout_id_2 <- row.names(data_rna[(data_rna[gene_2] == 0.0), ])
    dropout_id_3 <- row.names(data_rna[((data_rna[gene_1] == 0.0) & (data_rna[gene_2] == 0.0)), ])

    # Add dropout labels for plotting purposes
    data_rna$dropout <- NA
    data_rna[dropout_id_1, "dropout"]  <- gene_1
    data_rna[dropout_id_2, "dropout"]  <- gene_2
    data_rna[dropout_id_3, "dropout"]  <- "both"

    p <- ggplot(data_rna, aes(x=get(gene_1), y=get(gene_2), color=dropout)) +
            geom_point() +
            theme_classic() +
            labs(x=gene_1, y=gene_2) +
            NoLegend() +
            ggtitle(paste(gene_1, "vs", gene_2))
    plot_list[[idx]] <- ggplotGrob(p)
}

grid.arrange(grobs=plot_list, ncol=5)
```

# Taking a subset of surfaceome gene

For faster computation time only using surfaceome genes based upon average expression and number of cells expressed in.

```{r}
surfaceome_genes <- c(surfaceome_genes_all, genes)

# Get RNA expression counts
data_rna <- FetchData(seurat[["RNA"]], vars=surfaceome_genes)

# Number of cells a gene is expressed in
# Average expression of a gene
num_cells <- colSums(data_rna > 0)
avg_expression <- colMeans(data_rna)

# Put into one dataframe
df_genes <- data.frame(num_cells, avg_expression)
df_genes$gene <- row.names(df_genes)
df_genes <- df_genes %>% arrange(desc(avg_expression), desc(num_cells))

# Manually remove B2M because it is an outlier that skews the data
surfaceome_genes <- surfaceome_genes[surfaceome_genes != "B2M"]

# Select top and bottom genes based on expression and ncells
df_genes$use <- FALSE
df_genes[head(df_genes, 50)$gene, "use"] <- TRUE
df_genes[tail(df_genes, 50)$gene, "use"] <- TRUE
df_genes[genes, "use"] <- TRUE

# Subset surfaceome genes
surfaceome_genes <- df_genes[df_genes$use, "gene"]

ggplot(df_genes, aes(x=num_cells, y=avg_expression, color=use)) +
    geom_point() +
    scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red")) +
    theme_classic() +
    NoLegend()
```

# Normalization/Imputation


## Log normalization (rna)

```{r}
# # RAW COUNTS
# raw_rna <- FetchData(seurat[["RNA"]], vars=surfaceome_genes, layer="count")

# # LOG NORMALIZATION
# data_rna <- FetchData(seurat[["RNA"]], vars=surfaceome_genes, layer="data")
```

## SCT
```{r}
# # Re-run SCT 
# seurat <- seurat[surfaceome_genes, ]
# seurat <- SCTransform(seurat, return.only.var.genes=FALSE, verbose=T)

# # SCT
# data_sct <- FetchData(seurat[["SCT"]], vars=surfaceome_genes, layer="data")
```

## MAGIC
```{r}
# # MAGIC
# # Load conda environment
# myenvs <- reticulate::conda_list()
# envname <- myenvs$name[3]
# reticulate::use_condaenv(envname, required = TRUE)
# library(Rmagic)
# data_magic <- magic(raw_rna)$result
```

## SAVER
```{r}
# # SAVER
# # Generate SAVER predictions for those genes
# data_saver <- saver(t(raw_rna), estimates.only = TRUE, ncores=8)
# data_saver <- data.frame(t(data_saver))
```

## Store output

```{r}
# # Put all matrices in a list for easy access
# data_list <- list(data_rna, data_sct, data_magic, data_saver)
# names(data_list) <- c("rna", "sct", "magic", "saver")

# saveRDS(data_list, glue("{path_outs}/data_list.RDS"))

data_list <- readRDS(glue("{path_outs}/data_list.RDS"))
```

# Compare average gene expression

## All genes
```{r}
surfaceome_genes <- surfaceome_genes[!(surfaceome_genes %in% c("B2M", "EEF1A1"))]
methods <- c("rna", "sct", "magic", "saver")

df_avg <- data.frame(gene=surfaceome_genes)

for (method in methods) {
    data <- data_list[[method]]
    avg <- data.frame(colSums(data) / dim(data)[1])
    colnames(avg) <- method
    avg$gene <- row.names(avg)

   df_avg <-  merge(df_avg, avg, by="gene")
}

pheatmap(df_avg %>% column_to_rownames(var="gene"), scale="column", 
            cluster_col=TRUE, cluster_row=TRUE, show_rownames=FALSE)
```


## Marker genes

```{r}
plot_exp <- function(gene_1, gene_2) {

    plot_list <- list()
    methods <- c("rna", "sct", "magic", "saver")

    data <- data_list[["rna"]][c(gene_1, gene_2)]
    dropout_id_1 <- row.names(data[(data[gene_1] == 0.0), ])
    dropout_id_2 <- row.names(data[(data[gene_2] == 0.0), ])
    dropout_id_3 <- row.names(data[((data[gene_2] == 0.0) & (data[gene_1] == 0.0)), ])

    for (method in methods) {
        data <- data_list[[method]]

        data$dropout <- NA
        data[dropout_id_1, "dropout"]  <- gene_1
        data[dropout_id_2, "dropout"]  <- gene_2
        data[dropout_id_3, "dropout"]  <- "both"

        corr_method <- round(corr_list[[method]][gene_1, gene_2], 3)
        p_corr_method <- round(p_corr_list[[method]][gene_1, gene_2], 3)

        corr_cscore <- round(corr_list[["cscore"]][gene_1, gene_2], 3)
        p_corr_cscore <- round(p_corr_list[["cscore"]][gene_1, gene_2], 3)
        title <- glue("{method}: correlation = {corr_method}, pval={p_corr_method}\ncscore: correlation = {corr_cscore}, pval={p_corr_cscore}")
   

        p <- ggplot(data, aes(x=get(gene_1), y=get(gene_2), color=dropout)) +
                geom_point() +
                theme_classic() +
                xlab(gene_1) + ylab(gene_2) +
                ggtitle(title) +
                NoLegend()
                
        plot_list[[method]] <- p
    }

    p <- cowplot::plot_grid(plotlist=plot_list, ncol=2)
    return(p)
}
```

```{r}
df <- df_avg %>% subset(gene %in% genes)  %>% remove_rownames() %>% column_to_rownames(var="gene") 
pheatmap(df, scale="column", annotation_row=genes_label, annotation_colors=anno_colors,
            cluster_col=TRUE, cluster_row=TRUE)
```


# Correlation

## Spearman correlation

```{r}
run_spearman <- function(data_list, surfaceome_genes) {

    # Compute spearman correlation for each method (except CS-CORE)
    methods <- c("rna", "sct", "magic", "saver")
    p_corr_list <- list()
    corr_list <- list()

    # Unique combination of each surfaceome gene pair
    genes_comb <- data.frame(t(combn(surfaceome_genes, 2)))

    for (method in methods) {
        data <- data_list[[method]]

        gene_1_method <- c()
        gene_2_method <- c()
        corr_method <- c()
        p_method <- c()

        for (idx in 1:nrow(genes_comb)) {

            gene_1 <- genes_comb[idx, 1]
            gene_2 <- genes_comb[idx, 2]

            gene_1_method <- c(gene_1_method, gene_1)
            gene_2_method <- c(gene_2_method, gene_2)

            gene_1_exp <- data[[gene_1]]
            gene_2_exp <- data[[gene_2]]

            if ((length(gene_1_exp) == 0) | (length(gene_2_exp) == 0 | all(gene_1_exp == 0) | all(gene_2_exp == 0))) {
                # If a gene has no expression, set correlation = 0 and p-value = 1
                corr_val <- 0.0
                p_val <- 1.0
            } else {
                # Calculate spearman correlation and p-value otherwise
                tmp <- cor.test(gene_1_exp, gene_2_exp, method="spearman", exact=F)
                corr_val <- as.numeric(unname(tmp$estimate))
                p_val <- as.numeric(tmp$p.value)
            }

            corr_method <- c(corr_method, corr_val)
            p_method <- c(p_method, p_val)

        }

        # Data wrangling to put into matrix format
        corr <- data.frame(Var1=gene_1_method, Var2=gene_2_method, value=corr_method)
        corr_cp <- corr %>% rename(Var1=Var2, Var2=Var1)
        corr_cp <- corr_cp[c("Var1", "Var2", "value")]
        corr <- rbind(corr, corr_cp)
        corr <- reshape2::dcast(corr, Var2 ~ Var1)
        rownames(corr) <- corr$Var2
        corr <- subset(corr, select = -c(Var2))

        p_corr <- data.frame(Var1=gene_1_method, Var2=gene_2_method, value=p_method)
        p_corr_cp <- p_corr %>% rename(Var1=Var2, Var2=Var1)
        p_corr_cp <- p_corr_cp[c("Var1", "Var2", "value")]
        p_corr <- rbind(p_corr, p_corr_cp)
        p_corr <- reshape2::dcast(p_corr, Var2 ~ Var1)
        rownames(p_corr) <- p_corr$Var2
        p_corr <- subset(p_corr, select = -c(Var2))



        # Set the diagonal values
        # Correlation = 1, p-value = 1
        corr <- as.matrix(corr)
        diag(corr) <- 1

        p_corr <- as.matrix(p_corr)
        diag(p_corr) <- 1

        # Store results
        p_corr_list[[method]] <- p_corr
        corr_list[[method]] <- corr

    }

    results <- list()
    results[["corr"]] <- corr_list
    results[["p_val"]] <- p_corr_list

    return(results)

}
```

```{r}
# spearman_res <- run_spearman(data_list, surfaceome_genes)
# saveRDS(spearman_res, glue("{path_outs}/spearman_res.RDS"))

spearman_res <- readRDS(glue("{path_outs}/spearman_res.RDS"))
p_corr_list <- spearman_res[["p_val"]]
corr_list <- spearman_res[["corr"]]

```

## CS-CORE

```{r}
# Run CS-CORE
CSCORE_result <- CSCORE(seurat, genes = surfaceome_genes)

corr_list[["cscore"]] <- CSCORE_result$est
p_corr_list[["cscore"]] <- CSCORE_result$p_value
```

# BH adjusted matrix

```{r}
# Soft threshold parameters for WGCNA
power_list <- list()
power_list[["rna"]] <- 2
power_list[["sct"]] <- 1
power_list[["magic"]] <- 1
power_list[["saver"]] <- 4
power_list[["cscore"]] <- 1


run_wgcna <- function(mtx_coex, mtx_pval, genes, power) {
  # Obtain BH-adjusted p values
  p_matrix_BH <- matrix(0, length(genes), length(genes))
  p_matrix_BH[upper.tri(p_matrix_BH)] <- p.adjust(mtx_pval[upper.tri(mtx_pval)], method = "BH")
  p_matrix_BH <- p_matrix_BH + t(p_matrix_BH)

  # Set co-expression entires with BH-adjusted p-values greater than 0.05 to 0
  mtx_coex[p_matrix_BH > 0.05] <- 0

  adj <- WGCNA::adjacency.fromSimilarity(abs(mtx_coex), power = power)

  # Compute the topological overlap matrix
  TOM <- WGCNA::TOMsimilarity(adj)
  dissTOM <- 1 - TOM
  rownames(dissTOM) <- colnames(dissTOM) <- genes
  # Run hierarchical clustering as in the WGCNA workflow
  hclust_dist <- hclust(as.dist(dissTOM), method = "average") 
  memb <- dynamicTreeCut::cutreeDynamic(dendro = hclust_dist, 
                      distM = dissTOM, 
                      deepSplit = 2,
                      pamRespectsDendro = FALSE,
                      minClusterSize = 20)

  # For more instructions on how to tune the parameters in the WGCNA workflow,
  # please refer to https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/Rpackages/WGCNA/Tutorials/

  names(memb) <- genes
  memb_tab <- table(memb)
  module_list <- lapply(sort(unique(memb)), function(i_k) names(which(memb == i_k)))

  module_df <- data.frame(memb)
  module_df$memb <- as.character(module_df$memb)
  df <- module_df
  df <- tibble::rownames_to_column(df, "gene")
  colnames(df) <- c("gene", "module")

  # Return both the modules and BH-adjust matrix
  results <- list()
  results[["wgcna"]] <- df
  results[["bh_mtx"]] <- mtx_coex

  return(results)
}
```

```{r}
wgcna_list <- list()
bh_mtx_list <- list()

for (method in names(corr_list)) {
    corr <- corr_list[[method]]
    p_corr <- p_corr_list[[method]]
    power <- power_list[[method]]

    # Run WGCNA
    results <- run_wgcna(corr, p_corr, surfaceome_genes, power=power)
    wgcna_list[[method]] <- results$wgcna
    bh_mtx_list[[method]] <- results$bh_mtx
}
```

## MAGIC

### Raw spearman correlation results

```{r}
pheatmap(corr_list[["magic"]], show_rownames=FALSE, show_colnames=FALSE)
```

### Setting non-sig pairs to be 0 (BH adjusted matrix)

```{r}
pheatmap(bh_mtx_list[["magic"]], show_rownames=FALSE, show_colnames=FALSE)
```

### Marker genes
```{r}
pheatmap(bh_mtx_list[["magic"]][genes, genes], annotation_row=genes_label, annotation_colors=anno_colors, annotation_col=genes_label)
```

## SAVER

### Raw spearman correlation results

```{r}
pheatmap(corr_list[["saver"]], show_rownames=FALSE, show_colnames=FALSE)
```

### Setting non-sig pairs to be 0 (BH adjusted matrix)

```{r}
pheatmap(bh_mtx_list[["saver"]], show_rownames=FALSE, show_colnames=FALSE)
```


### Marker genes
```{r}
pheatmap(bh_mtx_list[["saver"]][genes, genes], annotation_row=genes_label, annotation_colors=anno_colors, annotation_col=genes_label)
```


## cscore

### Raw spearman correlation results

```{r}
pheatmap(corr_list[["cscore"]], show_rownames=FALSE, show_colnames=FALSE)
```

### Setting non-sig pairs to be 0 (BH adjusted matrix)

```{r}
pheatmap(bh_mtx_list[["cscore"]], show_rownames=FALSE, show_colnames=FALSE)
```

### Marker genes
```{r}
pheatmap(bh_mtx_list[["cscore"]][genes, genes], annotation_row=genes_label, annotation_colors=anno_colors, annotation_col=genes_label)
```


## Before filtering

First trying to select good filtration parameters.

```{r}
# For each method, calculate the average expression for each gene
comp_avg <- data.frame(gene=surfaceome_genes) 
for (method in names(data_list)) {
    df <- data_list[[method]]
    df <- data.frame(colMeans(df))
    colnames(df) <- c(method)
    row.names(df) <- str_replace(row.names(df), "\\.", "-")

    df$gene <- rownames(df)
    comp_avg <- merge(comp_avg, df)
}
comp_avg <- comp_avg %>% column_to_rownames(var="gene")

comp_n_cells <- data.frame(colSums(data_list[["rna"]] > 0))
colnames(comp_n_cells) <- "rna"

summary(comp_avg)
```


```{r}
min_exp <- 0.2
min_cells <- 40
min_perc <- 0.2

n_cells <- dim(data_list[["rna"]])[1]
if (min_perc * n_cells < 30) {
    min_perc <- 30 / n_cells
}
```

Histogram of the average expression for each gene.

```{r}
methods <- c("rna", "sct", "magic", "saver")
palette <- hue_pal()(length(methods))

plot_list <- list()
for (idx in 1:4) {
    method <- methods[idx]
    color <- palette[idx]

    p <- ggplot(comp_avg) +
        geom_histogram(aes(x=get(method)), fill=color, bins=50, alpha=0.5) +
        theme_classic() +
        labs(x="Average expression") +
        ggtitle(method) +
        geom_vline(xintercept = min_exp, color="black", size=1.5, linetype="dashed")

    plot_list[[idx]] <- p

}

cowplot::plot_grid(plotlist=plot_list, ncol=2)
```

Histogram of the number of cells each gene is expressed in based upon the log-normalized expression.

```{r}
ggplot(comp_n_cells) +
    geom_histogram(aes(x=rna), bins=50, alpha=0.5, color="black", fill=palette[1]) +
    theme_classic() +
    labs(x="Number of cells") +
    ggtitle("RNA number of cells") +
   geom_vline(xintercept = min_cells, color="black", size=1.5, linetype="dashed")
```

Histogram of the percentage of cells each gene is expressed in based upon the log-normalized expression (rna).


```{r}
comp_perc_cells <- comp_n_cells / n_cells

ggplot(comp_perc_cells) +
    geom_histogram(aes(x=rna), bins=50, alpha=0.5, color="black", fill=palette[1]) +
    theme_classic() +
    labs(x="Percentage of cells") +
    ggtitle("RNA percentage of cells") +
    geom_vline(xintercept = min_perc, color="black", size=1.5, linetype="dashed")
```

Now I compare each method against one another by plotting the correlation scores for each gene pair. Any pairs that would be filtered by the thresholds (listed below) are highlighted in red on the scatterplots. Only one gene in the pair needs to fail this metric to be filtered out.


- average expression < `r min_exp`
- number of cells expressed in > `r min_cells`
- percentag of cells expressed in > `r min_perc`


For CS-CORE which does not provide a count matrix, I am using the values from whichever method it is being compared against (RNA, SCT, MAGIC, SAVER).

```{r}
# Co-expression/correlation scores for each method
comp_corr <- reshape2::melt(corr_list[["rna"]]) %>% rename(rna=value)
for (method in c("sct", "magic", "saver", "cscore")) {
    df <- corr_list[[method]]
    df <- reshape2::melt(df)
    colnames(df) <- c("Var1", "Var2", method)

    comp_corr <- merge(comp_corr, df)
}

final_pairs <- comp_corr
final_pairs <- final_pairs %>% subset(Var1 != Var2)

# Get genes filtered by min cells
filter_genes <- comp_n_cells %>% subset(rna < min_cells)
filter_genes <- rownames(filter_genes)

final_pairs <- final_pairs %>% mutate(filter = ((Var1 %in% filter_genes) | (Var2 %in% filter_genes)))
```


```{r fig.height=12.5, fig.width=12.5}
methods <- names(bh_mtx_list)
plot_list <- list()
idx <- 1

for (method_1 in methods) {
    for (method_2 in methods) {

        if (method_1 == "cscore") {comp_avg$cscore = comp_avg[[method_2]]}
        if (method_2 == "cscore") {comp_avg$cscore = comp_avg[[method_1]]}
        
        # Genes to filter by min expression
        tmp <- comp_avg %>% subset((get(method_1) < min_exp) & (get(method_2) < min_exp))
        filter_genes <- row.names(tmp)

        df <- final_pairs
        df[(df$Var1 %in% filter_genes) | (df$Var2 %in% filter_genes), "filter"] <- TRUE
        df <- df %>% arrange(filter)

        p <- ggplot(df) +
            geom_point(aes(x=df[[method_1]], y=df[[method_2]], color=df[["filter"]])) +
            scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red")) +
            theme_classic() +
            NoLegend() +
            labs(x=method_1, y=method_2) +
            ggtitle(paste(method_1, "vs", method_2)) +
            ylim(-1, 1) +
            xlim(-1, 1)


        plot_list[[idx]] <- ggplotGrob(p)

        idx <- idx + 1
    }
}
grid.arrange(grobs=plot_list, ncol=5)
```

## After filtration

```{r fig.height=12.5, fig.width=12.5}
methods <- names(bh_mtx_list)
plot_list <- list()
idx <- 1

for (method_1 in methods) {
    for (method_2 in methods) {

        if (method_1 == "cscore") {comp_avg$cscore = comp_avg[[method_2]]}
        if (method_2 == "cscore") {comp_avg$cscore = comp_avg[[method_1]]}
        
        # Genes to filter by min expression
        tmp <- comp_avg %>% subset((get(method_1) < min_exp) & (get(method_2) < min_exp))
        filter_genes <- row.names(tmp)

        df <- final_pairs
        df[(df$Var1 %in% filter_genes) | (df$Var2 %in% filter_genes), "filter"] <- TRUE
        df <- df %>% arrange(filter)

        df <- df %>% subset(filter == FALSE)

        p <- ggplot(df) +
            geom_point(aes(x=df[[method_1]], y=df[[method_2]], color=df[["filter"]])) +
            scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red")) +
            theme_classic() +
            NoLegend() +
            labs(x=method_1, y=method_2) +
            ggtitle(paste(method_1, "vs", method_2)) +
            ylim(-1, 1) +
            xlim(-1, 1) +
            geom_abline(slope=1, intercept=0, color="red")


        plot_list[[idx]] <- ggplotGrob(p)

        idx <- idx + 1
    }
}
grid.arrange(grobs=plot_list, ncol=5)
```

# Comparing specific methods

## MAGIC

Magic tends to always give high correlation scores


### vs saver
```{r}
method_1 <- "saver"
method_2 <- "magic"

if (method_1 == "cscore") {comp_avg$cscore = comp_avg[[method_2]]}
if (method_2 == "cscore") {comp_avg$cscore = comp_avg[[method_1]]}

# Genes to filter by min expression
tmp <- comp_avg %>% subset((get(method_1) < min_exp) & (get(method_2) < min_exp))
filter_genes <- row.names(tmp)

df <- final_pairs
df[(df$Var1 %in% filter_genes) | (df$Var2 %in% filter_genes), "filter"] <- TRUE
df <- df %>% arrange(filter)

df <- df %>% subset(filter == FALSE)

p <- ggplot(df) +
    geom_point(aes(x=get(method_1), y=get(method_2), color=filter)) +
    scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red")) +
    theme_classic() +
    NoLegend() +
    labs(x=method_1, y=method_2) +
    ggtitle(paste(method_1, "vs", method_2)) +
    ylim(-1, 1) +
    xlim(-1, 1) +
    geom_abline(slope=1, intercept=0, color="red")
p
```

### vs cscore

```{r}
method_1 <- "cscore"
method_2 <- "magic"

if (method_1 == "cscore") {comp_avg$cscore = comp_avg[[method_2]]}
if (method_2 == "cscore") {comp_avg$cscore = comp_avg[[method_1]]}

# Genes to filter by min expression
tmp <- comp_avg %>% subset((get(method_1) < min_exp) & (get(method_2) < min_exp))
filter_genes <- row.names(tmp)

df <- final_pairs
df[(df$Var1 %in% filter_genes) | (df$Var2 %in% filter_genes), "filter"] <- TRUE
df <- df %>% arrange(filter)

df <- df %>% subset(filter == FALSE)

p <- ggplot(df) +
    geom_point(aes(x=get(method_1), y=get(method_2), color=filter)) +
    scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red")) +
    theme_classic() +
    NoLegend() +
    labs(x=method_1, y=method_2) +
    ggtitle(paste(method_1, "vs", method_2)) +
    ylim(-1, 1) +
    xlim(-1, 1) +
    geom_abline(slope=1, intercept=0, color="red")
p
```

## CS-CORE vs SAVER

Tend to agree, but there are exceptions

```{r}
method_1 <- "cscore"
method_2 <- "saver"

if (method_1 == "cscore") {comp_avg$cscore = comp_avg[[method_2]]}
if (method_2 == "cscore") {comp_avg$cscore = comp_avg[[method_1]]}

# Genes to filter by min expression
tmp <- comp_avg %>% subset((get(method_1) < min_exp) & (get(method_2) < min_exp))
filter_genes <- row.names(tmp)

df <- final_pairs
df[(df$Var1 %in% filter_genes) | (df$Var2 %in% filter_genes), "filter"] <- TRUE
df <- df %>% arrange(filter)

df <- df %>% subset(filter == FALSE)

p <- ggplot(df) +
    geom_point(aes(x=get(method_1), y=get(method_2), color=filter)) +
    scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red")) +
    theme_classic() +
    NoLegend() +
    labs(x=method_1, y=method_2) +
    ggtitle(paste(method_1, "vs", method_2)) +
    ylim(-1, 1) +
    xlim(-1, 1) +
    geom_abline(slope=1, intercept=0, color="red")
p
```

# Why do cscore and saver disagree?

## Differences in significance

recall we set correlations to 0 when the gene pairs are not significant

if we undo this, the relationship looks like this:

```{r}
comp_corr <- reshape2::melt(corr_list[["rna"]]) %>% rename(rna=value)
for (method in c("sct", "magic", "saver", "cscore")) {
    df <- corr_list[[method]]
    df <- reshape2::melt(df)
    colnames(df) <- c("Var1", "Var2", method)

    comp_corr <- merge(comp_corr, df)
}

method_1 <- "cscore"
method_2 <- "saver"

if (method_1 == "cscore") {comp_avg$cscore = comp_avg[[method_2]]}
if (method_2 == "cscore") {comp_avg$cscore = comp_avg[[method_1]]}

df <- comp_corr %>% subset(Var1 != Var2)
# Get genes filtered by min cells
filter_genes <- comp_n_cells %>% subset(rna < min_cells)
filter_genes <- rownames(filter_genes)

df <- df %>% mutate(filter = ((Var1 %in% filter_genes) | (Var2 %in% filter_genes)))


# Genes to filter by min expression
tmp <- comp_avg %>% subset((get(method_1) < min_exp) & (get(method_2) < min_exp))
filter_genes <- row.names(tmp)

df[(df$Var1 %in% filter_genes) | (df$Var2 %in% filter_genes), "filter"] <- TRUE
df <- df %>% arrange(filter)

df <- df %>% subset(filter == FALSE)
```

```{r}
p <- ggplot(df) +
    geom_point(aes(x=get(method_1), y=get(method_2), color=filter)) +
    scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red")) +
    theme_classic() +
    NoLegend() +
    labs(x=method_1, y=method_2) +
    ggtitle(paste(method_1, "vs", method_2)) +
    ylim(-1, 1) +
    xlim(-1, 1) +
    geom_abline(slope=1, intercept=0, color="red")
p
```


## Big differences in expression values of the genes

Large changes in average expression of gene 1 and gene 2 - not considered significant

```{r}
gene_1 <- "CCL5"
gene_2 <- "CD8A"

comp_avg[c(gene_1, gene_2), , drop=FALSE]
```

```{r fig.height=8, fig.width=8}
plot_exp(gene_1, gene_2)
```


# Only keep significant gene pairs where cscore and saver agree

```{r}
method_1 <- "cscore"
method_2 <- "saver"

if (method_1 == "cscore") {comp_avg$cscore = comp_avg[[method_2]]}
if (method_2 == "cscore") {comp_avg$cscore = comp_avg[[method_1]]}

# Genes to filter by min expression
tmp <- comp_avg %>% subset((get(method_1) < min_exp) & (get(method_2) < min_exp))
filter_genes <- row.names(tmp)

df <- final_pairs
df[(df$Var1 %in% filter_genes) | (df$Var2 %in% filter_genes), "filter"] <- TRUE
df <- df %>% arrange(filter)

df <- df %>% subset(filter == FALSE)

p <- ggplot(df) +
    geom_point(aes(x=get(method_1), y=get(method_2), color=filter)) +
    scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red")) +
    theme_classic() +
    NoLegend() +
    labs(x=method_1, y=method_2) +
    ggtitle(paste(method_1, "vs", method_2)) +
    ylim(-1, 1) +
    xlim(-1, 1) +
    geom_abline(slope=1, intercept=0, color="red")
p
```


Becomes 

```{r}
df <- df %>% subset(saver!=0 & cscore!=0)

p <- ggplot(df) +
    geom_point(aes(x=get(method_1), y=get(method_2), color=filter)) +
    scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red")) +
    theme_classic() +
    NoLegend() +
    labs(x=method_1, y=method_2) +
    ggtitle(paste(method_1, "vs", method_2)) +
    ylim(-1, 1) +
    xlim(-1, 1) +
    geom_abline(slope=1, intercept=0, color="red")
p
```

With the final table of genes like so:

```{r}
df <- df %>% arrange(desc(saver), desc(cscore))
df <- df[!duplicated(cbind(t(apply(df[1:2], 1, sort)))), ]
df
```

## Example

Using the `gene_lookup()` function, we can supply the first gene pair listed to look at the summary information of the pair.

```{r fig.height=8, fig.width=8}
idx <- 1
gene_1 <- as.character(df[idx, "Var1"])
gene_2 <- as.character(df[idx, "Var2"])
plot_exp(gene_1, gene_2)
```


# Plot PNG

```{r fig.height=8, fig.width=8}
idx <- 1
gene_1 <- "CCL5"
gene_2 <- "CST7"

plot_list <- list()
methods <- c("rna", "saver")

data <- data_list[["rna"]][c(gene_1, gene_2)]
dropout_id_1 <- row.names(data[(data[gene_1] == 0.0), ])
dropout_id_2 <- row.names(data[(data[gene_2] == 0.0), ])
dropout_id_3 <- row.names(data[((data[gene_2] == 0.0) & (data[gene_1] == 0.0)), ])

for (method in methods) {
    data <- data_list[[method]]

    data$dropout <- NA
    data[dropout_id_1, "dropout"]  <- gene_1
    data[dropout_id_2, "dropout"]  <- gene_2
    data[dropout_id_3, "dropout"]  <- "both"

    corr_method <- round(corr_list[[method]][gene_1, gene_2], 3)
    p_corr_method <- round(p_corr_list[[method]][gene_1, gene_2], 3)

    corr_cscore <- round(corr_list[["cscore"]][gene_1, gene_2], 3)
    p_corr_cscore <- round(p_corr_list[["cscore"]][gene_1, gene_2], 3)


    if (method == "rna") {title <- "Log-normalized"}
    else {title <- "SAVER imputed"}

    p <- ggplot(data, aes(x=get(gene_1), y=get(gene_2), color=dropout)) +
            geom_point() +
            theme_classic() +
            xlab(gefne_1) + ylab(gene_2) +
            ggtitle(title) +
            NoLegend() +
            theme(plot.title = element_text(size=rel(2)), 
            axis.title = element_text(size=rel(1.5)))

            
    plot_list[[method]] <- p
}

png("../../plots/ccl5_cst7_saver.png", , res=200, width=10,height=5,units="in")

cowplot::plot_grid(plotlist=plot_list, ncol=2) 

dev.off()


```


```{r fig.height=8, fig.width=8}
idx <- 1
gene_1 <- "CCL5"
gene_2 <- "CST7"

plot_list <- list()
methods <- c("rna", "magic")

data <- data_list[["rna"]][c(gene_1, gene_2)]
dropout_id_1 <- row.names(data[(data[gene_1] == 0.0), ])
dropout_id_2 <- row.names(data[(data[gene_2] == 0.0), ])
dropout_id_3 <- row.names(data[((data[gene_2] == 0.0) & (data[gene_1] == 0.0)), ])

for (method in methods) {
    data <- data_list[[method]]

    data$dropout <- NA
    data[dropout_id_1, "dropout"]  <- gene_1
    data[dropout_id_2, "dropout"]  <- gene_2
    data[dropout_id_3, "dropout"]  <- "both"

    corr_method <- round(corr_list[[method]][gene_1, gene_2], 3)
    p_corr_method <- round(p_corr_list[[method]][gene_1, gene_2], 3)

    corr_cscore <- round(corr_list[["cscore"]][gene_1, gene_2], 3)
    p_corr_cscore <- round(p_corr_list[["cscore"]][gene_1, gene_2], 3)


    if (method == "rna") {title <- "Log-normalized"}
    else {title <- "MAGIC imputed"}

    p <- ggplot(data, aes(x=get(gene_1), y=get(gene_2), color=dropout)) +
            geom_point() +
            theme_classic() +
            xlab(gene_1) + ylab(gene_2) +
            ggtitle(title) +
            NoLegend() +
            theme(plot.title = element_text(size=rel(2)), 
            axis.title = element_text(size=rel(1.5)))

            
    plot_list[[method]] <- p
}

png("../../plots/ccl5_cst7_magic.png", , res=200, width=10,height=5,units="in")

cowplot::plot_grid(plotlist=plot_list, ncol=2) 

dev.off()


```



```{r fig.height=8, fig.width=8}
idx <- 1
gene_1 <- "CCL5"
gene_2 <- "CST7"

plot_list <- list()
methods <- c("rna", "sct")

data <- data_list[["rna"]][c(gene_1, gene_2)]
dropout_id_1 <- row.names(data[(data[gene_1] == 0.0), ])
dropout_id_2 <- row.names(data[(data[gene_2] == 0.0), ])
dropout_id_3 <- row.names(data[((data[gene_2] == 0.0) & (data[gene_1] == 0.0)), ])

for (method in methods) {
    data <- data_list[[method]]

    data$dropout <- NA
    data[dropout_id_1, "dropout"]  <- gene_1
    data[dropout_id_2, "dropout"]  <- gene_2
    data[dropout_id_3, "dropout"]  <- "both"

    corr_method <- round(corr_list[[method]][gene_1, gene_2], 3)
    p_corr_method <- round(p_corr_list[[method]][gene_1, gene_2], 3)

    corr_cscore <- round(corr_list[["cscore"]][gene_1, gene_2], 3)
    p_corr_cscore <- round(p_corr_list[["cscore"]][gene_1, gene_2], 3)


    if (method == "rna") {title <- "Log-normalized"}
    else {title <- "SCTransformed"}

    p <- ggplot(data, aes(x=get(gene_1), y=get(gene_2), color=dropout)) +
            geom_point() +
            theme_classic() +
            xlab(gene_1) + ylab(gene_2) +
            ggtitle(title) +
            NoLegend() +
            theme(plot.title = element_text(size=rel(2)), 
            axis.title = element_text(size=rel(1.5)))

            
    plot_list[[method]] <- p
}

png("../../plots/ccl5_cst7_sct.png", , res=200, width=10,height=5,units="in")

cowplot::plot_grid(plotlist=plot_list, ncol=2) 

dev.off()


```


```{r fig.height=8, fig.width=8}
idx <- 1
gene_1 <- "CCL5"
gene_2 <- "CST7"

plot_list <- list()
methods <- c("rna")

data <- data_list[["rna"]][c(gene_1, gene_2)]
dropout_id_1 <- row.names(data[(data[gene_1] == 0.0), ])
dropout_id_2 <- row.names(data[(data[gene_2] == 0.0), ])
dropout_id_3 <- row.names(data[((data[gene_2] == 0.0) & (data[gene_1] == 0.0)), ])

for (method in methods) {
    data <- data_list[[method]]

    data$dropout <- NA
    data[dropout_id_1, "dropout"]  <- gene_1
    data[dropout_id_2, "dropout"]  <- gene_2
    data[dropout_id_3, "dropout"]  <- "both"

    corr_method <- round(corr_list[[method]][gene_1, gene_2], 3)
    p_corr_method <- round(p_corr_list[[method]][gene_1, gene_2], 3)

    corr_cscore <- round(corr_list[["cscore"]][gene_1, gene_2], 3)
    p_corr_cscore <- round(p_corr_list[["cscore"]][gene_1, gene_2], 3)


    if (method == "rna") {title <- "Log-normalized"}
    else {title <- "MAGIC imputed"}

    p <- ggplot(data, aes(x=get(gene_1), y=get(gene_2), color=dropout)) +
            geom_point() +
            theme_classic() +
            xlab(gene_1) + ylab(gene_2) +
            ggtitle(title) +
            NoLegend() +
            theme(plot.title = element_text(size=rel(2)), 
            axis.title = element_text(size=rel(1.5)))

            
    plot_list[[method]] <- p
}

png("../../plots/ccl5_cst7_rna.png", res=200, width=5,height=5,units="in")

cowplot::plot_grid(plotlist=plot_list, ncol=1) 

dev.off()
```



```{r}
df_avg_impute <- data.frame(surfaceome_genes) %>% rename(gene = surfaceome_genes)

# For each method, calculate the average expression for each gene
for (method in names(data_list)) {
    df <- data_list[[method]]
    df <- data.frame(colMeans(df))
    colnames(df) <- c(method)
    df$gene <- rownames(df)

    df_avg_impute <- left_join(df_avg_impute, df)
}

df_avg_impute <- df_avg_impute %>% column_to_rownames(var="gene")
pheatmap(df_avg_impute,  cluster_rows = TRUE, cluster_cols = TRUE, show_rownames = FALSE, scale="column")

```


```{r}
# Co-expression/correlation scores for each method
comp_corr <- reshape2::melt(corr_list[["rna"]]) %>% rename(rna=value)
for (method in c("sct", "magic", "saver", "cscore")) {
    df <- corr_list[[method]]
    df <- reshape2::melt(df)
    colnames(df) <- c("Var1", "Var2", method)

    comp_corr <- merge(comp_corr, df)
}

final_pairs <- comp_corr
final_pairs <- final_pairs %>% subset(Var1 != Var2)

# Get genes filtered by min cells
filter_genes <- comp_n_cells %>% subset(rna < min_cells)
filter_genes <- rownames(filter_genes)

final_pairs <- final_pairs %>% mutate(filter = ((Var1 %in% filter_genes) | (Var2 %in% filter_genes)))

methods <- names(bh_mtx_list)
plot_list <- list()
idx <- 1

for (method_1 in methods) {
    for (method_2 in methods) {

        if (method_1 == "cscore") {comp_avg$cscore = comp_avg[[method_2]]}
        if (method_2 == "cscore") {comp_avg$cscore = comp_avg[[method_1]]}
        
        # Genes to filter by min expression
        tmp <- comp_avg %>% subset((get(method_1) < min_exp) & (get(method_2) < min_exp))
        filter_genes <- row.names(tmp)

        df <- final_pairs
        df[(df$Var1 %in% filter_genes) | (df$Var2 %in% filter_genes), "filter"] <- TRUE
        df <- df %>% arrange(filter)

        p <- ggplot(df) +
            geom_point(aes(x=df[[method_1]], y=df[[method_2]], color=df[["filter"]])) +
            scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red")) +
            theme_classic() +
            NoLegend() +
            labs(x=method_1, y=method_2) +
            ggtitle(paste(method_1, "vs", method_2)) +
            ylim(-1, 1) +
            xlim(-1, 1) +
            theme(plot.title = element_text(size=rel(2)), )


        plot_list[[idx]] <- ggplotGrob(p)

        idx <- idx + 1
    }
}

cols <- c(1, 2, 7, 3, 8, 13, 4, 9, 14, 19, 5, 10, 15, 20, 25)
plot_list_sub <- plot_list[cols]

m <- matrix(NA, 5, 5)
m[upper.tri(m, diag = T)] <- 1:15


png("../../plots/compare_before_filt.png",  res=200, width=15,height=15,units="in")
grid.arrange(grobs=plot_list_sub, layout_matrix = m)
dev.off()
```


```{r}
# Co-expression/correlation scores for each method
comp_corr <- reshape2::melt(corr_list[["rna"]]) %>% rename(rna=value)
for (method in c("sct", "magic", "saver", "cscore")) {
    df <- corr_list[[method]]
    df <- reshape2::melt(df)
    colnames(df) <- c("Var1", "Var2", method)

    comp_corr <- merge(comp_corr, df)
}

final_pairs <- comp_corr
final_pairs <- final_pairs %>% subset(Var1 != Var2)

# Get genes filtered by min cells
filter_genes <- comp_n_cells %>% subset(rna < min_cells)
filter_genes <- rownames(filter_genes)

final_pairs <- final_pairs %>% mutate(filter = ((Var1 %in% filter_genes) | (Var2 %in% filter_genes)))

methods <- names(bh_mtx_list)
plot_list <- list()
idx <- 1

for (method_1 in methods) {
    for (method_2 in methods) {

        if (method_1 == "cscore") {comp_avg$cscore = comp_avg[[method_2]]}
        if (method_2 == "cscore") {comp_avg$cscore = comp_avg[[method_1]]}
        
        # Genes to filter by min expression
        tmp <- comp_avg %>% subset((get(method_1) < min_exp) & (get(method_2) < min_exp))
        filter_genes <- row.names(tmp)

        df <- final_pairs
        df[(df$Var1 %in% filter_genes) | (df$Var2 %in% filter_genes), "filter"] <- TRUE
        df <- df %>% subset(filter == FALSE)

        p <- ggplot(df) +
            geom_point(aes(x=df[[method_1]], y=df[[method_2]], color=df[["filter"]])) +
            scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red")) +
            theme_classic() +
            NoLegend() +
            labs(x=method_1, y=method_2) +
            ggtitle(paste(method_1, "vs", method_2)) +
            ylim(-1, 1) +
            xlim(-1, 1) +
            theme(plot.title = element_text(size=rel(2)), )


        plot_list[[idx]] <- ggplotGrob(p)

        idx <- idx + 1
    }
}

cols <- c(1, 2, 7, 3, 8, 13, 4, 9, 14, 19, 5, 10, 15, 20, 25)
plot_list_sub <- plot_list[cols]

m <- matrix(NA, 5, 5)
m[upper.tri(m, diag = T)] <- 1:15


png("../../plots/compare_after_filt.png",  res=200, width=15,height=15,units="in")
grid.arrange(grobs=plot_list_sub, layout_matrix = m)
dev.off()
```


```{r}
plot_list <- list()

method_1 <- "magic"
method_2 <- "cscore"

if (method_1 == "cscore") {comp_avg$cscore = comp_avg[[method_2]]}
if (method_2 == "cscore") {comp_avg$cscore = comp_avg[[method_1]]}

# Genes to filter by min expression
tmp <- comp_avg %>% subset((get(method_1) < min_exp) & (get(method_2) < min_exp))
filter_genes <- row.names(tmp)

df <- final_pairs
df[(df$Var1 %in% filter_genes) | (df$Var2 %in% filter_genes), "filter"] <- TRUE
df <- df %>% arrange(filter)

df <- df %>% subset(filter == FALSE)

p <- ggplot(df) +
    geom_point(aes(x=get(method_1), y=get(method_2), color=filter)) +
    scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red")) +
    theme_classic() +
    NoLegend() +
    labs(x=method_1, y=method_2) +
    ggtitle("MAGIC vs CS-CORE") +
    ylim(-1, 1) +
    xlim(-1, 1) +
    geom_abline(slope=1, intercept=0, color="red") +
    theme(plot.title = element_text(size=rel(2)), )
plot_list[[1]] <- p



method_1 <- "magic"
method_2 <- "saver"

if (method_1 == "cscore") {comp_avg$cscore = comp_avg[[method_2]]}
if (method_2 == "cscore") {comp_avg$cscore = comp_avg[[method_1]]}

# Genes to filter by min expression
tmp <- comp_avg %>% subset((get(method_1) < min_exp) & (get(method_2) < min_exp))
filter_genes <- row.names(tmp)

df <- final_pairs
df[(df$Var1 %in% filter_genes) | (df$Var2 %in% filter_genes), "filter"] <- TRUE
df <- df %>% arrange(filter)

df <- df %>% subset(filter == FALSE)

p <- ggplot(df) +
    geom_point(aes(x=get(method_1), y=get(method_2), color=filter)) +
    scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red")) +
    theme_classic() +
    NoLegend() +
    labs(x=method_1, y=method_2) +
    ggtitle("MAGIC vs SAVER") +
    ylim(-1, 1) +
    xlim(-1, 1) +
    geom_abline(slope=1, intercept=0, color="red") +
    theme(plot.title = element_text(size=rel(2)), )
plot_list[[2]] <- p

png("../../plots/compare_magic.png",  res=200, width=5,height=10,units="in")

cowplot::plot_grid(plotlist=plot_list, ncol=1)

dev.off()
```

```{r}
# Co-expression/correlation scores for each method
comp_corr <- reshape2::melt(corr_list[["rna"]]) %>% rename(rna=value)
for (method in c("sct", "magic", "saver", "cscore")) {
    df <- corr_list[[method]]
    df <- reshape2::melt(df)
    colnames(df) <- c("Var1", "Var2", method)

    comp_corr <- merge(comp_corr, df)
}

final_pairs <- comp_corr
final_pairs <- final_pairs %>% subset(Var1 != Var2)

# Get genes filtered by min cells
filter_genes <- comp_n_cells %>% subset(rna < min_cells)
filter_genes <- rownames(filter_genes)

final_pairs <- final_pairs %>% mutate(filter = ((Var1 %in% filter_genes) | (Var2 %in% filter_genes)))


method_1 <- "cscore"
method_2 <- "saver"

if (method_1 == "cscore") {comp_avg$cscore = comp_avg[[method_2]]}
if (method_2 == "cscore") {comp_avg$cscore = comp_avg[[method_1]]}

# Genes to filter by min expression
tmp <- comp_avg %>% subset((get(method_1) < min_exp) & (get(method_2) < min_exp))
filter_genes <- row.names(tmp)


df <- final_pairs
df[(df$Var1 %in% filter_genes) | (df$Var2 %in% filter_genes), "filter"] <- TRUE
df <- df %>% arrange(filter)

df <- df %>% subset(filter == FALSE)


# Co-expression/correlation scores for each method
comp_p <- reshape2::melt(p_corr_list[["rna"]]) %>% rename(rna=value)
for (method in c("sct", "magic", "saver", "cscore")) {
    df_ <- p_corr_list[[method]]
    df_ <- reshape2::melt(df_)
    colnames(df_) <- c("Var1", "Var2", method)

    comp_p <- merge(comp_p, df_)
}
comp_p <- comp_p %>% select(Var1, Var2, saver, cscore) %>% rename(p_saver=saver, p_cscore=cscore)

df <- left_join(df, comp_p, join_by(Var1==Var1, Var2==Var2))


df$sig <- FALSE
df[df$p_saver < 0.05, "sig"] <- "saver"
df[df$p_cscore < 0.05, "sig"] <- "cscore"
df[(df$p_cscore < 0.05) & (df$p_saver < 0.05), "sig"] <- "both"
df[(df$p_cscore > 0.05) & (df$p_saver > 0.05), "sig"] <- "neither"



png("../../plots/compare_saver_cscore_sig.png",  res=200, width=6,height=5,units="in")
ggplot(df) +
    geom_point(aes(x=get(method_1), y=get(method_2), color=sig)) +
    theme_classic() +
    labs(x=method_1, y=method_2) +
    ggtitle("CS-CORE vs SAVER") +
    ylim(-1, 1) +
    xlim(-1, 1) +
    geom_abline(slope=1, intercept=0, color="red") +
    theme(plot.title = element_text(size=rel(2)))

dev.off()
```



```{r fig.height=8, fig.width=8}
idx <- 1
gene_1 <- "CCL5"
gene_2 <- "CST7"

plot_list <- list()
methods <- c("saver")

data <- data_list[["rna"]][c(gene_1, gene_2)]
dropout_id_1 <- row.names(data[(data[gene_1] == 0.0), ])
dropout_id_2 <- row.names(data[(data[gene_2] == 0.0), ])
dropout_id_3 <- row.names(data[((data[gene_2] == 0.0) & (data[gene_1] == 0.0)), ])

for (method in methods) {
    data <- data_list[[method]]

    data$dropout <- NA
    data[dropout_id_1, "dropout"]  <- gene_1
    data[dropout_id_2, "dropout"]  <- gene_2
    data[dropout_id_3, "dropout"]  <- "both"

    corr_method <- round(corr_list[[method]][gene_1, gene_2], 3)
    p_corr_method <- round(p_corr_list[[method]][gene_1, gene_2], 3)

    corr_cscore <- round(corr_list[["cscore"]][gene_1, gene_2], 3)
    p_corr_cscore <- round(p_corr_list[["cscore"]][gene_1, gene_2], 3)


    if (method == "rna") {title <- "Log-normalized"}
    else {title <- "SAVER imputed expression"}

    p <- ggplot(data, aes(x=get(gene_1), y=get(gene_2), color=dropout)) +
            geom_point() +
            theme_classic() +
            xlab(gene_1) + ylab(gene_2) +
            ggtitle(title) +
            NoLegend() +
            theme(plot.title = element_text(size=rel(2)), 
            axis.title = element_text(size=rel(1.5)))

            
    plot_list[[method]] <- p
}

png("../../plots/cd8a_ccl5_saver.png", , res=200, width=5,height=5,units="in")

cowplot::plot_grid(plotlist=plot_list, ncol=1) 

dev.off()


```

```{r}
gene_1 <- "CIB1"
gene_2 <- "FYN"

plot_list <- list()
methods <- c("saver")

data <- data_list[["rna"]][c(gene_1, gene_2)]
dropout_id_1 <- row.names(data[(data[gene_1] == 0.0), ])
dropout_id_2 <- row.names(data[(data[gene_2] == 0.0), ])
dropout_id_3 <- row.names(data[((data[gene_2] == 0.0) & (data[gene_1] == 0.0)), ])

for (method in methods) {
    data <- data_list[[method]]

    data$dropout <- NA
    data[dropout_id_1, "dropout"]  <- gene_1
    data[dropout_id_2, "dropout"]  <- gene_2
    data[dropout_id_3, "dropout"]  <- "both"

    corr_method <- round(corr_list[[method]][gene_1, gene_2], 3)
    p_corr_method <- round(p_corr_list[[method]][gene_1, gene_2], 3)

    corr_cscore <- round(corr_list[["cscore"]][gene_1, gene_2], 3)
    p_corr_cscore <- round(p_corr_list[["cscore"]][gene_1, gene_2], 3)


    if (method == "rna") {title <- "Log-normalized expression"}
    if (method == "sct") {title <- "SCTransform expression"}
    if (method == "magic") {title <- "Magic imputed expression"}
    if (method == "saver") {title <- "SAVER imputed expression"}

    p <- ggplot(data, aes(x=get(gene_1), y=get(gene_2), color=dropout)) +
            geom_point() +
            theme_classic() +
            xlab(gene_1) + ylab(gene_2) +
            ggtitle(title) +
            NoLegend() +
            theme(plot.title = element_text(size=rel(2)), 
            axis.title = element_text(size=rel(1.5)))

            
    plot_list[[method]] <- p
}

png("../../plots/CIB1_fyn_saver.png", , res=200, width=5,height=5,units="in")

cowplot::plot_grid(plotlist=plot_list, ncol=1) 

dev.off()

```


```{r}
test <- df %>% subset(sig == "saver") %>% arrange(desc(saver))
```

```{r}
comp_corr %>% subset(Var1==gene_1 & Var2==gene_2)
```


```{r}
gene_1 <- "CST7"
gene_2 <- "CCL5"

exp <- data.frame(gene_1=data_list[["saver"]][gene_1], gene_2=data_list[["saver"]][gene_2])

info <- data.frame(avg=colMeans(exp), stdev=apply(exp, 2, sd)  )
info

```


```{r}
gene_1 <- "PLSCR1"
gene_2 <- "LY6E"

plot_list <- list()
methods <- c("rna", "sct", "magic", "saver")

data <- data_list[["rna"]][c(gene_1, gene_2)]
dropout_id_1 <- row.names(data[(data[gene_1] == 0.0), ])
dropout_id_2 <- row.names(data[(data[gene_2] == 0.0), ])
dropout_id_3 <- row.names(data[((data[gene_2] == 0.0) & (data[gene_1] == 0.0)), ])

for (method in methods) {
    data <- data_list[[method]]

    data$dropout <- NA
    data[dropout_id_1, "dropout"]  <- gene_1
    data[dropout_id_2, "dropout"]  <- gene_2
    data[dropout_id_3, "dropout"]  <- "both"

    corr_method <- round(corr_list[[method]][gene_1, gene_2], 3)
    p_corr_method <- round(p_corr_list[[method]][gene_1, gene_2], 3)

    corr_cscore <- round(corr_list[["cscore"]][gene_1, gene_2], 3)
    p_corr_cscore <- round(p_corr_list[["cscore"]][gene_1, gene_2], 3)


    if (method == "rna") {title <- "Log-normalized expression"}
    if (method == "sct") {title <- "SCTransform expression"}
    if (method == "magic") {title <- "Magic imputed expression"}
    if (method == "saver") {title <- "SAVER imputed expression"}

    p <- ggplot(data, aes(x=get(gene_1), y=get(gene_2), color=dropout)) +
            geom_point() +
            theme_classic() +
            xlab(gene_1) + ylab(gene_2) +
            ggtitle(title) +
            NoLegend() +
            theme(plot.title = element_text(size=rel(2)), 
            axis.title = element_text(size=rel(1.5)))

            
    plot_list[[method]] <- p
}

png(glue("../../plots/{gene_1}_{gene_2}.png"), res=200, width=10,height=10,units="in")

cowplot::plot_grid(plotlist=plot_list, ncol=2) 

dev.off()

```



```{r}
# Co-expression/correlation scores for each method
comp_corr <- reshape2::melt(corr_list[["rna"]]) %>% rename(rna=value)
for (method in c("sct", "magic", "saver", "cscore")) {
    df <- corr_list[[method]]
    df <- reshape2::melt(df)
    colnames(df) <- c("Var1", "Var2", method)

    comp_corr <- merge(comp_corr, df)
}

final_pairs <- comp_corr
final_pairs <- final_pairs %>% subset(Var1 != Var2)

# Get genes filtered by min cells
filter_genes <- comp_n_cells %>% subset(rna < min_cells)
filter_genes <- rownames(filter_genes)

final_pairs <- final_pairs %>% mutate(filter = ((Var1 %in% filter_genes) | (Var2 %in% filter_genes)))


method_1 <- "cscore"
method_2 <- "saver"

if (method_1 == "cscore") {comp_avg$cscore = comp_avg[[method_2]]}
if (method_2 == "cscore") {comp_avg$cscore = comp_avg[[method_1]]}

# Genes to filter by min expression
tmp <- comp_avg %>% subset((get(method_1) < min_exp) & (get(method_2) < min_exp))
filter_genes <- row.names(tmp)


df <- final_pairs
df[(df$Var1 %in% filter_genes) | (df$Var2 %in% filter_genes), "filter"] <- TRUE
df <- df %>% arrange(filter)

df <- df %>% subset(filter == FALSE)


# Co-expression/correlation scores for each method
comp_p <- reshape2::melt(p_corr_list[["rna"]]) %>% rename(rna=value)
for (method in c("sct", "magic", "saver", "cscore")) {
    df_ <- p_corr_list[[method]]
    df_ <- reshape2::melt(df_)
    colnames(df_) <- c("Var1", "Var2", method)

    comp_p <- merge(comp_p, df_)
}
comp_p <- comp_p %>% select(Var1, Var2, saver, cscore) %>% rename(p_saver=saver, p_cscore=cscore)

df <- left_join(df, comp_p, join_by(Var1==Var1, Var2==Var2))


df$sig <- FALSE
df[df$p_saver < 0.05, "sig"] <- "saver"
df[df$p_cscore < 0.05, "sig"] <- "cscore"
df[(df$p_cscore < 0.05) & (df$p_saver < 0.05), "sig"] <- "both"
df[(df$p_cscore > 0.05) & (df$p_saver > 0.05), "sig"] <- "neither"

df <- df %>% subset(sig == "both")

png("../../plots/compare_saver_cscore_sig_filt.png",  res=200, width=5,height=5,units="in")
ggplot(df) +
    geom_point(aes(x=get(method_1), y=get(method_2)), color="black") +
    theme_classic() +
    labs(x=method_1, y=method_2) +
    ggtitle("CS-CORE vs SAVER") +
    ylim(-1, 1) +
    xlim(-1, 1) +
    theme(plot.title = element_text(size=rel(2)))

dev.off()
```

```{r}
df <- df <- df[!duplicated(cbind(t(apply(df[1:2], 1, sort)))), ]
df %>% subset(saver > 0.5 & cscore > 0.5) %>% arrange(desc(cscore)) %>% head()
```


## AVerages and gene selection

```{r}
# Load dataset
# remotes::install_github("mojaveazure/seurat-disk", quiet=F)
library(SeuratDisk)
load(bzfile("../data/seurat_integrated.RData.bz2"))

seurat <- seurat_integrated
DefaultAssay(seurat) <- "RNA"
seurat[["SCT"]] <- NULL
seurat[["integrated"]] <- NULL

rm(seurat_integrated)

Idents(object = seurat) <- "integrated_snn_res.0.8"
seurat <- RenameIdents(object = seurat, 
                               "0" = "Naive or memory CD4+ T cells",
                               "1" = "CD14+ monocytes",
                               "2" = "Activated T cells",
                               "3" = "CD14+ monocytes",
                               "4" = "Stressed cells / Unknown",
                               "5" = "CD8+ T cells",
                               "6" = "Naive or memory CD4+ T cells",
                               "7" = "B cells",
                               "8" = "NK cells",
                               "9" = "CD8+ T cells",
                               "10" = "FCGR3A+ monocytes",
                               "11" = "B cells",
                               "12" = "NK cells",
                               "13" = "B cells",
                               "14" = "Conventional dendritic cells",
                               "15" = "Megakaryocytes",
			       "16" = "Plasmacytoid dendritic cells")

seurat$celltype <- Idents(seurat)



seurat <- subset(seurat, subset = (celltype == "CD8+ T cells"))

# Removing genes that have 0 counts across all cells of the celltype
counts <- seurat[["RNA"]]@counts
genes.use <- rowSums(counts) > 0.0
genes.use <- names(genes.use[genes.use])
seurat <- seurat[genes.use, ]
rm(counts)
```



```{r}
# Re-run SCT with hvgs set as surfaceome genes to make sure that they get normalized
# VariableFeatures(seurat) <- surfaceome_genes
# seurat <- SCTransform(seurat, return.only.var.genes=FALSE)

# Get the updated list of surfaceome genes for these cells
surfaceome_genes <- read.csv("../data/surfaceome_genes.csv")$x

# Get the updated list of surfaceome genes found in the dataset
surfaceome_genes_all <- surfaceome_genes[surfaceome_genes %in% genes.use]
surfaceome_genes_all <- c(surfaceome_genes_all, genes)
```



Additionally, I am going to select the top 100 genes (within the surfaceome gene list) based upon average expression and number of cells it is expressed in. Similarly, I am taking the bottom 50. So in total, we will be running this pipeline around 150 genes + selected marker genes (which are labelled in red). I am subsetting the genes for the sake of saving time for this report as we define best practices.

```{r fig.height=5, fig.width=8}
data_rna <- FetchData(seurat[["RNA"]], vars=surfaceome_genes_all, layer="data")

# Number of cells a gene is expressed in
# Average expression of a gene
num_cells <- colSums(data_sct > 0)
avg_expression <- colMeans(data_sct)

df_genes <- data.frame(num_cells=num_cells, avg_expression=avg_expression)
df_genes$gene <- row.names(df_genes)
df_genes <- df_genes %>% arrange(num_cells, avg_expression)

# Select top 100 and bottom 50 genes based on expression and ncells
# Select top and bottom genes based on expression and ncells
df_genes$use <- FALSE
df_genes[head(df_genes, 50)$gene, "use"] <- TRUE
df_genes[tail(df_genes, 50)$gene, "use"] <- TRUE
df_genes[df_genes[450:460, ]$gene, "use"] <- TRUE
df_genes[genes, "use"] <- TRUE

df_genes <- df_genes %>% arrange(use)

df_genes %>% tail()

ggplot(df_genes, aes(x=num_cells, y=avg_expression, color=use)) +
  geom_point() +
  theme_classic() +
  geom_text(aes(label=ifelse(gene %in% genes, as.character(gene), '')), hjust=-0.1, vjust=0.1, color="black") +
  scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red")) +
  NoLegend()
```


```{r}
# Number of cells a gene is expressed in
# Average expression of a gene
num_cells <- colSums(data_rna > 0)
avg_expression <- colMeans(data_rna)

df_genes <- data.frame(num_cells, avg_expression)
df_genes$gene <- row.names(df_genes)
df_genes <- df_genes %>% arrange(desc(avg_expression), desc(num_cells))

# Select top and bottom genes based on expression and ncells
surfaceome_genes <- c(head(df_genes, 50)$gene, tail(df_genes, 50)$gene)
surfaceome_genes <- unique(c(surfaceome_genes, genes))

# Manually remove B2M because it is an outlier that skews the data
surfaceome_genes <- surfaceome_genes[surfaceome_genes != "B2M"]

```








```{r}
# MAGIC
# Load conda environment
myenvs <- reticulate::conda_list()
envname <- myenvs$name[3]
reticulate::use_condaenv(envname, required = TRUE)
library(Rmagic)
data_magic <- magic(raw_rna)$result

# SAVER
# Generate SAVER predictions for those genes
data_saver <- saver(t(raw_rna), estimates.only = TRUE, ncores=8)
data_saver <- data.frame(t(data_saver))

# # LOG NORMALIZATION
data_rna <- FetchData(seurat[["RNA"]], vars=surfaceome_genes, layer="data")

```