---
title: "Merck - CS-CORE pbmc_multimodal"
author: "Noor Sohail"
output:
   html_document:
      code_folding: hide
      df_print: paged
      highlights: pygments
      number_sections: true
      self_contained: true
      theme: default
      toc: true
      toc_float:
         collapsed: true
         smooth_scroll: true
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
# Turn off Warnings and other console output messages from the whole document
```

```{r}
#' Create sub-chunks for plots
#' Plotting function to generate plots of a specific dimension when html is knitted
#'
#' taken from: https://stackoverflow.com/questions/15365829/dynamic-height-and-width-for-knitr-plots
#'
#' @param pl a plot object
#' @param fig.height figure height
#' @param fig.width figure width
#' @param chunk_name name of the chunk
#'
#' @author Andreas Scharmueller \email{andschar@@protonmail.com}
#'
subchunkify = function(pl,
                       fig.height = 7,
                       fig.width = 5,
                       chunk_name = 'plot') {
  pl_deparsed = paste0(deparse(function() {
    pl
  }), collapse = '')
  
  sub_chunk = paste0(
    "```{r ",
    chunk_name,
    ", fig.height=",
    fig.height,
    ", fig.width=",
    fig.width,
    ", dpi=72",
    ", echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}",
    "\n(",
    pl_deparsed,
    ")()",
    "\n```"
  )
  
  cat(knitr::knit(
    text = knitr::knit_expand(text = sub_chunk),
    quiet = TRUE
  ))
}
```

```{r, cache=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(glue)
library(Seurat)
library(scales)
library(ggh4x)
library(gridExtra)
library(pheatmap)
library(devtools)

library(CSCORE)
library(WGCNA)

ggplot2::theme_set(theme_light(base_size = 11))
opts_chunk[["set"]](
    cache = FALSE,
    dev = c("png", "pdf"),
    error = TRUE,
    highlight = TRUE,
    message = FALSE,
    prompt = FALSE,
    tidy = FALSE,
    warning = FALSE)

axis <- ggh4x::guide_axis_truncated(
  trunc_lower = unit(0, "npc"),
  trunc_upper = unit(2, "cm")
)
```

# Project details

``` {r}
set.seed(2023)

name_project <- "Co-expression of genes: pbmc_multimodal"
name_pi <- "Merck"

name_file <- gsub(" ", "_", name_project)
name_file <- gsub("-", "_", name_file)
```


<b>Dataset background</b>

This is a sc-RNA-Seq data that James merged from individual sorted pbmc subset populations that he downloaded from here:
https://www.10xgenomics.com/resources/datasets?query=&page=1&configure%5Bfacets%5D%5B0[…]s%5D%5B1%5D=pipeline.version&configure%5BhitsPerPage%5D=500

A brief overview of what the dataset looks like:

```{r}
# Load dataset
# remotes::install_github("mojaveazure/seurat-disk", quiet=F)
library(SeuratDisk)
seurat <- LoadH5Seurat("../data/pbmc_multimodal.h5seurat")
```

```{r fig.height=6}
seurat$celltype <- seurat$celltype.l1
seurat@meta.data %>%
        ggplot() +
        geom_bar(aes(
            x = celltype,
            fill=celltype),
            stat = "count", color = "black") +
        theme_classic() +
        NoLegend() +
        xlab("Celltype") +
        ylab("Number of Cells") +
        ggtitle(glue("{name_project}\nCelltypes")) +
        theme(plot.title = element_text(hjust = 0.5)) +
        geom_text(aes(x = celltype, label = after_stat(count)), stat = "count", vjust = -0.5) +
        theme(axis.text.x = element_text(angle = 45, hjust=1))
```

```{r fig.height=5, fig.width=7}
Idents(seurat) <- "celltype"
p <- DimPlot(seurat, reduction = "umap") +
    xlab("UMAP1") +
    ylab("UMAP2") +
    ggtitle(glue("{name_project}\nCelltypes")) +
    theme(plot.title = element_text(hjust = 0.5)) +
    theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(),
        axis.text.y=element_blank(), axis.ticks.y=element_blank(),
        axis.title=element_text(size=8, face="bold")) +
    guides(x = axis, y = axis) +
    theme(axis.line = element_line(arrow = arrow(type = "closed"), size = 1),
        axis.title = element_text(hjust = 0))
LabelClusters(p, id = "ident",  fontface = "bold", size = 3, bg.colour = "white", bg.r = .2, force = 0)
```

# Surfaceome genes
```{r}
# surfaceome <- read.csv("../data/surfaceome.csv")

# library('biomaRt')
# mart <- useDataset("hsapiens_gene_ensembl", useMart("ensembl"))
# gene_IDs <- getBM(filters= "ensembl_gene_id", attributes= c("ensembl_gene_id","hgnc_symbol"),
#               values = surfaceome$ENSEMBL.Gene.ID, mart= mart)

# surfaceome_genes <- left_join(x = surfaceome, y = gene_IDs, by = c("ENSEMBL.Gene.ID" = "ensembl_gene_id"))
# surfaceome_genes <- surfaceome_genes$hgnc_symbol

# surfaceome_genes <- rownames(seurat$RNA)[rownames(seurat$RNA) %in% surfaceome_genes]

# write.csv(surfaceome_genes, "../data/surfaceome_genes.csv")
surfaceome_genes <- read.csv("../data/surfaceome_genes.csv")$x
as.data.frame(surfaceome_genes)
```

# Installing CS-CORE

The installation process for the tool is very straightforward and I had no issues getting it working. 

```{r}
# ## Load devtools for installing R packages from Github
# library(devtools)
# ## Install CS-CORE from Github
# install_github("ChangSuBiostats/CS-CORE")
```

# Running CS-CORE

In the vignette for [CS-CORE](https://changsubiostats.github.io/CS-CORE/articles/CSCORE.html) they subset their dataset by a singular celltype and then input their correlation matrix into [WGCNA](https://cran.r-project.org/web/packages/WGCNA/index.html) in order to determine modules of genes. 

```{r fig.height=6}
celltypes <- sort(unique((seurat$celltype)))

Idents(seurat) <- "celltype"
seurat@meta.data %>% 
    ggplot() +
    geom_bar(aes(
        x = Idents(seurat),
        fill = Idents(seurat)), stat = "count", color = "black", show.legend = FALSE) +
    theme_classic() +
    xlab("Celltype") +
    ylab("Number of Cells") +
    ggtitle(glue("{name_project}\nManual celltypes")) +
    theme(plot.title = element_text(hjust = 0.5)) +
    geom_text(aes(x = Idents(seurat), label = after_stat(count)), stat = "count", vjust = -0.5) +
    theme(axis.text.x = element_text(angle = 45, hjust=1))
```

# Co-expression by celltypes

```{r}
# This dataset only came with ADT and SCT assays
# The raw counts were stored in the "counts" slot of the SCT assay
# Creating another RNA assay with the raw counts as CSCORE requires it
raw_rna <- GetAssayData(object =  seurat[['SCT']], layer = 'counts')
seurat[["RNA"]] <- CreateAssayObject(counts = raw_rna )
seurat <- NormalizeData(seurat, assay="RNA")
```

```{r results="asis", echo=FALSE}
for (ct in celltypes) {

    cat(paste0("## ", ct, "\n\n"))
    cat(paste0("To save computation time, I remove genes that have 0 counts in the ", ct, " subsetted dataset.\n"))
    seurat_sub <- subset(seurat, subset=(celltype == ct))

    # Removing genes that have 0 counts across all cells of the celltype
    counts <- seurat_sub[["RNA"]]@counts
    genes.use <- rowSums(counts) > 0
    genes.use <- names(genes.use[genes.use])
    seurat_sub <- seurat_sub[genes.use, ]

    # Surfaceome genes in the dataset
    genes <- surfaceome_genes[surfaceome_genes %in% row.names(seurat_sub)]
    n_genes <- length(genes)

    cat(paste0("After this gene filtration, we have ", n_genes, " genes from the surfaceome to use as input for CS-CORE.\n\n"))

    # Run CS-CORE
    CSCORE_result <- CSCORE(seurat_sub, genes = genes)



    # cat("\n\n### Co-expression heatmap\n\n")
    # cat(paste0("Here we can see the estimated correlation estimates within the ", ct, " cells.\n\n"))
    # p <- pheatmap(CSCORE_result$est, show_colnames = FALSE, show_rownames = FALSE, silent=TRUE)
    # subchunkify(p, n_genes/10, n_genes/10, glue("{ct}-coexp"))
    # cat("\n\n")



    cat("### BH-adjusted matrix\n\n")
    cat("The vignette then has us replace gene pairs that are no significant to be 0 within the estimated co-expression matrix. This step is necessary in order to feed the CS-CORE co-expression estimated into WGCNA to determine gene modules as it originally takes in a pearson correlation matrix.\n\n") 
    cat("In essence, we are setting the co-expression values of non-significant gene pairs to be 0 - which in the context of a pearson correlation matrix would mean that the pairs are not correlated and therefore will not have co-expression.\n\n")
    # cat("The modified matrix now looks like this, with the 0'd values:\n\n")

    # Obtain CS-CORE co-expression estimates
    CSCORE_coexp <- CSCORE_result$est

    # Obtain BH-adjusted p values
    CSCORE_p <- CSCORE_result$p_value
    p_matrix_BH <- matrix(0, length(genes), length(genes))
    p_matrix_BH[upper.tri(p_matrix_BH)] <- p.adjust(CSCORE_p[upper.tri(CSCORE_p)], method = "BH")
    p_matrix_BH <- p_matrix_BH + t(p_matrix_BH)

    # Set co-expression entires with BH-adjusted p-values greater than 0.05 to 0
    CSCORE_coexp[p_matrix_BH > 0.05] <- 0

    # p <- pheatmap(CSCORE_coexp, show_colnames = FALSE, show_rownames = FALSE, silent=TRUE)
    # subchunkify(p, n_genes/10, n_genes/10, glue("{ct}-coexp_zero"))
    # cat("\n\n")

    
    genes.use <- rowSums(CSCORE_coexp) > 1.0
    genes.use <- names(genes.use[genes.use])
    n_genes <- length(genes.use)
    cat(glue("Now we want to remove genes that have no significant correlated expression with any genes. In doing so, we are now working with {n_genes} for this analysis.\n\n"))
    CSCORE_coexp <- CSCORE_coexp[genes.use, genes.use]
    # p <- pheatmap(CSCORE_coexp, show_colnames = FALSE, show_rownames = FALSE, silent=TRUE)
    # subchunkify(p, 10, 10, glue("{ct}-coexp_zero_subset"))
    cat("\n\n")


    cat("### WGCNA\n\n")
    cat("We then feed this 0'd co-expression matrix into WGCNA.\n\n")
    # Compute the adjacency matrix based on the co-expression matrix
    adj <- WGCNA::adjacency.fromSimilarity(abs(CSCORE_coexp), power = 1)
    # Compute the topological overlap matrix
    TOM <- WGCNA::TOMsimilarity(adj)
    dissTOM <- 1-TOM
    rownames(dissTOM) <- colnames(dissTOM) <- genes.use
    # Run hierarchical clustering as in the WGCNA workflow
    hclust_dist <- hclust(as.dist(dissTOM), method = "average") 
    memb <- dynamicTreeCut::cutreeDynamic(dendro = hclust_dist, 
                        distM = dissTOM, 
                        deepSplit = 2,
                        pamRespectsDendro = FALSE,
                        minClusterSize = 10)
    # For more instructions on how to tune the parameters in the WGCNA workflow,
    # please refer to https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/Rpackages/WGCNA/Tutorials/

    names(memb) <- genes.use
    memb_tab <- table(memb)
    module_list <- lapply(sort(unique(memb)), function(i_k) names(which(memb == i_k)))

    module_df <- data.frame(memb)
    module_df$memb <- as.character(module_df$memb)
    df <- module_df
    df <- tibble::rownames_to_column(df, "gene")
    colnames(df) <- c("gene", "module")

    cat(knitr::knit_print(df))
    cat("\n\n\n")
    cat("We are now given a list of modules with groups of genes that are co-expressed together in our dataset.\n\n")



    cat("### CS-CORE correlation estimates for modules\n")
    cat("Please note that this is the same correlation matrix in the section 'Co-expression heatmap', but now with annotations for the modules those genes belong to. There are no overlaps of genes among the modules.\n\n")

    module_df$genes <- row.names(module_df)
    anno <- module_df[, c("memb"), drop=F]
    # Co-expression matrix of only genes in the module
    coex_est <- CSCORE_result$est[module_df$genes, module_df$genes]
    p <- pheatmap(coex_est, show_colnames=F, 
            show_rownames=F, annotation_col=anno, annotation_row=anno, silent=T)
    subchunkify(p, 10, 11, glue("{ct}-module_coex"))
    cat("\n\n")

    # Get coexpression pairs estimate
    coex_df <- reshape2::melt(coex_est)
    colnames(coex_df) <- c("Var1", "Var2", "coex_est")

    # Get p-value of coexpression pairs
    p_est <- CSCORE_result$p_value[module_df$genes, module_df$genes]
    p_est <- reshape2::melt(p_est)
    coex_df$p_value <- p_est$value

    # Remove pairs with coexpression = 0
    # Remove pairs of the same gene (correlation will automatically be 1)
    coex_df <- coex_df %>% 
            filter(coex_est != 0) %>% 
            filter(Var1 != Var2)

    # Remove non-significant correlations
    coex_df <- coex_df %>%
        filter(p_value <= 0.05) %>%
        arrange(p_value)


    # Add module information
    coex_df$Var1_module <- coex_df$Var1
    coex_df$Var2_module <- coex_df$Var2
    for (idx in 1:nrow(module_df)) {
        module <- module_df[idx, ]$memb
        gene <- module_df[idx, ]$genes

        coex_df$Var1_module <- str_replace(coex_df$Var1_module, paste0(gene, "$"), module)
        coex_df$Var2_module <- str_replace(coex_df$Var2_module, paste0(gene, "$"), module)
    }

    # Remove pairs that are the same but ordered differently
    # Ex. (Var1, Var2) = (Var2, Var1)
    pairs <- c()
    for (idx in 1:nrow(coex_df)) {
        pair <- sort(c(coex_df[idx, ]$Var1, coex_df[idx, ]$Var2))
        pair <- paste0(pair[1], "+", pair[2])
        pairs <- c(pairs, c(pair))
    }
    coex_df$pairs <- pairs
    coex_df <- coex_df[!duplicated(coex_df$pairs), ]

    # Save output
    write.csv(coex_df, glue("02.1_results-cs_core/{ct}.csv"))



    cat("### Top co-expressed pairs\n\n")
    cat(glue("I have removed gene pairs that are not significant (0.05) and sorted the values by p-value to more easily identify interesting gene pairs within {ct}. I have also written out the results as a csv file before this filtration in case there are other thresholds we want to explore.\n\n"))
    # Remove non-significant correlations
    coex_df <- coex_df %>%
        filter(p_value <= 0.05) %>%
        arrange(p_value)
    cat(knitr::knit_print(coex_df))
    
    cat("\n\n\n")



    cat("### Scatterplot top pair\n\n")
    cat(glue("As a sanity check, I wanted to visualize the expression of the top correlated pair within the {ct} population. I selected the top significant gene pair that had more than 100 cells expressing both genes.\n\n"))

    n_cells <- 0
    idx <- 1
    while (n_cells < 100) {
        # Gene name of top co-expressed genes
        gene_1 <- as.character(coex_df[idx, "Var1"])
        gene_2 <- as.character(coex_df[idx, "Var2"])
        # Counts for genes
        counts <- data.frame(t(data.frame(seurat_sub$RNA@data[c(gene_1, gene_2), ])))
        colnames(counts) <- c(gene_1, gene_2)

        n_zero <- dim(counts)[1]
        n_cells <- dim(counts)[1]
        # Remove cells with 0 expression for both genes
        counts_zero <- counts[rowSums(counts) > 0, ]
        n_zero <- n_zero - dim(counts_zero)[1]

        n_cells <- dim(counts_zero)[1]
        idx <- idx + 1
    }

    # Scatterplot
    scatter <- ggplot(counts) +
        geom_point(aes(x=get(gene_1), y=get(gene_2))) +
        scale_fill_continuous(type = "viridis") +
        theme_classic() +
        xlab(gene_1) +
        ylab(gene_2) +
        geom_abline(slope=1, intercept = 0, color="red")

    cat(glue("As the scatterplot can make it difficult to see how many cells overlap at the same values, I have taken the same information displayed as a density plot. To create a better scale, I removed cells that have 0 expression for both genes ({n_zero} cells).\n\n"))

    # Density scatterplot
    density <- ggplot(counts_zero) +
        geom_bin2d(aes(x=get(gene_1), y=get(gene_2)), bins=70) +
        scale_fill_continuous(type = "viridis") +
        theme_classic() +
        xlab(gene_1) +
        ylab(gene_2) +
        geom_abline(slope=1, intercept = 0, color="red")

    p <- cowplot::plot_grid(scatter, density, ncol=2)
    subchunkify(p, 6, 13, glue("{ct}-scatter"))
    cat("\n\n")
}
```

# Next

This is the FACs sorted datset - compare the genes they were sorted on.