---
title: "Protein expression for pbmc_multimodal"
author: "Noor Sohail"
output:
   html_document:
      code_folding: hide
      df_print: paged
      highlights: pygments
      number_sections: true
      self_contained: true
      theme: default
      toc: true
      toc_float:
         collapsed: true
         smooth_scroll: true
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
# Turn off Warnings and other console output messages from the whole document
```

```{r}
#' Create sub-chunks for plots
#' Plotting function to generate plots of a specific dimension when html is knitted
#'
#' taken from: https://stackoverflow.com/questions/15365829/dynamic-height-and-width-for-knitr-plots
#'
#' @param pl a plot object
#' @param fig.height figure height
#' @param fig.width figure width
#' @param chunk_name name of the chunk
#'
#' @author Andreas Scharmueller \email{andschar@@protonmail.com}
#'
subchunkify = function(pl,
                       fig.height = 7,
                       fig.width = 5,
                       chunk_name = 'plot') {
  pl_deparsed = paste0(deparse(function() {
    pl
  }), collapse = '')
  
  sub_chunk = paste0(
    "```{r ",
    chunk_name,
    ", fig.height=",
    fig.height,
    ", fig.width=",
    fig.width,
    ", dpi=72",
    ", echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}",
    "\n(",
    pl_deparsed,
    ")()",
    "\n```"
  )
  
  cat(knitr::knit(
    text = knitr::knit_expand(text = sub_chunk),
    quiet = TRUE
  ))
}
```

```{r, cache=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(glue)
library(Seurat)
library(scales)
library(ggh4x)
library(gridExtra)
library(pheatmap)
library(devtools)

# library(CSCORE)
# library(WGCNA)
library(ggpubr)
library(ggvenn)

ggplot2::theme_set(theme_light(base_size = 11))
opts_chunk[["set"]](
    cache = FALSE,
    dev = c("png", "pdf"),
    error = TRUE,
    highlight = TRUE,
    message = FALSE,
    prompt = FALSE,
    tidy = FALSE,
    warning = FALSE)

axis <- ggh4x::guide_axis_truncated(
  trunc_lower = unit(0, "npc"),
  trunc_upper = unit(2, "cm")
)
```

# Project details

``` {r}
set.seed(2023)

name_project <- "Co-expression of genes: pbmc_multimodal"
name_pi <- "Merck"

name_file <- gsub(" ", "_", name_project)
name_file <- gsub("-", "_", name_file)
```


<b>Dataset background</b>
A brief overview of what the dataset looks like:

```{r}
# Load dataset
# remotes::install_github("mojaveazure/seurat-disk", quiet=F)
library(SeuratDisk)
seurat <- LoadH5Seurat("../data/pbmc_multimodal.h5seurat")
```

```{r}
# This dataset only came with ADT and SCT assays
# The raw counts were stored in the "counts" slot of the SCT assay
# Creating another RNA assay with the raw counts as CSCORE requires it
raw_rna <- GetAssayData(object =  seurat[['SCT']], layer = 'counts')
seurat[["RNA"]] <- CreateAssayObject(counts = raw_rna )
seurat <- NormalizeData(seurat, assay="RNA")

DefaultAssay(seurat) <- "RNA"
```

```{r fig.height=6}
seurat$celltype <- seurat$celltype.l1
seurat@meta.data %>%
        ggplot() +
        geom_bar(aes(
            x = celltype,
            fill=celltype),
            stat = "count", color = "black") +
        theme_classic() +
        NoLegend() +
        xlab("Celltype") +
        ylab("Number of Cells") +
        ggtitle(glue("{name_project}\nCelltypes")) +
        theme(plot.title = element_text(hjust = 0.5)) +
        geom_text(aes(x = celltype, label = after_stat(count)), stat = "count", vjust = -0.5) +
        theme(axis.text.x = element_text(angle = 45, hjust=1))
```

```{r fig.height=5, fig.width=7}
Idents(seurat) <- "celltype"
p <- DimPlot(seurat, reduction = "umap") +
    xlab("UMAP1") +
    ylab("UMAP2") +
    ggtitle(glue("{name_project}\nCelltypes")) +
    theme(plot.title = element_text(hjust = 0.5)) +
    theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(),
        axis.text.y=element_blank(), axis.ticks.y=element_blank(),
        axis.title=element_text(size=8, face="bold")) +
    guides(x = axis, y = axis) +
    theme(axis.line = element_line(arrow = arrow(type = "closed"), size = 1),
        axis.title = element_text(hjust = 0))
LabelClusters(p, id = "ident",  fontface = "bold", size = 3, bg.colour = "white", bg.r = .2, force = 0)
```

# Protein -> RNA naming

I am going to investigate the dropout rate more closely. 

To start, I'm going to look at genes that have the same name as the protein in the dataset for a first pass look.

```{r}
# Protein names
genes_adt <- sort(row.names(seurat[["ADT"]]))

genes <- genes_adt[genes_adt %in% row.names(seurat[["RNA"]])]
genes
```

# RNA vs Protein expression

Here I am comparing the normalized protein expression after applying a log10 transormation.

```{r}
# Log scale the ADT counts
counts_adt <- FetchData(seurat[["ADT"]], vars=genes)
counts_adt <- log10(counts_adt + 0.01) - log10(0.01)

# Log scale the RNA normalized counts
counts_rna <- FetchData(seurat[["RNA"]], vars=genes)
counts_rna <- log10(counts_rna + 0.01) - log10(0.01)

# Celltype information for each cell
celltype <- unname(seurat$celltype)
```

```{r results="asis", echo=FALSE}
for (gene in genes) {
    cat(glue("## {gene}\n\n"))
    gene <- gene
    protein <- gene


    cat("### All cells\n")
    title <- paste0("RNA ", gene, " vs Protein ", protein)
    scatter <- ggplot() +
        geom_point(aes(x=counts_rna[[gene]], y=counts_adt[[protein]], color=celltype)) +
        theme_classic() +
        xlab(glue("log10(RNA {gene})")) +
        ylab(glue("log10(Protein {protein})")) +
        geom_abline(slope=1, intercept = 0, color="red") +
        stat_cor(aes(x=counts_rna[[gene]], y=counts_adt[[protein]]), method = "pearson", color="blue") +
        ggtitle(title)

    density <- ggplot() +
        geom_bin2d(aes(x=counts_rna[[gene]], y=counts_adt[[protein]]), bins=100) +
        scale_fill_continuous(type = "viridis") +
        theme_classic() +
        xlab(glue("log10(RNA {gene})")) +
        ylab(glue("log10(Protein {protein})")) +
        geom_abline(slope=1, intercept = 0, color="red")

    p <- cowplot::plot_grid(scatter, density, ncol=2)
    subchunkify(p, 6, 12, glue("{gene}_{protein}-scatter"))
    cat("\n\n")


    cat("### Venn Diagram\n")
    cat("Here I am showing a venn diagram for the number/propotion of cells that 0 expression for each protein/gene pair. We are interested in the intersection, where we have cells that have expression in both RNA and Protein.\n")
    bool_rna <- counts_rna[[gene]] != 0.0
    bool_adt <- counts_adt[[gene]] != 0.0
    df <- data.frame(bool_rna, bool_adt)
    colnames(df) <- c("RNA", "Protein")
    p <- ggvenn(
        df, 
        stroke_size = 0.5, 
        set_name_size = 4) +
        ggtitle(glue("RNA {gene} + Protein {protein}: Zero expression"))
    subchunkify(p, 5, 5, glue("{gene}_{protein}-venn"))
    cat("\n\n")



    cat("### Remove cells with 0s\n")
    cat("Now I remove any cell that has 0 expression from either RNA or Protein.\n")
    title <- paste0("Non-zero: RNA ", gene, " vs Protein ", protein)
    df <- data.frame(counts_rna[[gene]], counts_adt[[protein]], celltype)
    colnames(df) <- c("RNA", "Protein", "celltype")
    df <- df[df$RNA != 0,]
    df <- df[df$Protein != 0,]

    p <- ggplot(df) +
        geom_bar(stat="count", aes(x=celltype, fill=celltype), color="black") +
        theme_classic() +
        ggtitle(title)
    subchunkify(p, 6, 12, glue("{gene}_{protein}-bar_zero"))
    cat("\n\n")

    scatter <- ggplot() +
        geom_point(aes(x=df$RNA, y=df$Protein, color=df$celltype)) +
        theme_classic() +
        xlab(glue("log10(RNA {gene})")) +
        ylab(glue("log10(Protein {protein})")) +
        geom_abline(slope=1, intercept = 0, color="red") +
        stat_cor(aes(x=df$RNA, y=df$Protein), method = "pearson", color="blue") +
        ggtitle(title)


    density <- ggplot() +
        geom_bin2d(aes(x=df$RNA, y=df$Protein), bins=100) +
        scale_fill_continuous(type = "viridis") +
        theme_classic() +
        xlab(glue("log10(RNA {gene})")) +
        ylab(glue("log10(Protein {protein})")) +
        geom_abline(slope=1, intercept = 0, color="red")

    p <- cowplot::plot_grid(scatter, density, ncol=2)
    subchunkify(p, 6, 12, glue("{gene}_{protein}-scatter_zero"))
    cat("\n\n")
}
```
