---
title: "Protein expression for pbmc_multimodal"
author: "Noor Sohail"
output:
   html_document:
      code_folding: hide
      df_print: paged
      highlights: pygments
      number_sections: true
      self_contained: true
      theme: default
      toc: true
      toc_float:
         collapsed: true
         smooth_scroll: true
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
# Turn off Warnings and other console output messages from the whole document
```

```{r, cache=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(glue)
library(Seurat)
library(scales)
library(ggh4x)
library(gridExtra)
library(pheatmap)
library(devtools)

# library(CSCORE)
# library(WGCNA)
library(ggpubr)
library(ggvenn)

ggplot2::theme_set(theme_light(base_size = 11))
opts_chunk[["set"]](
    cache = FALSE,
    dev = c("png", "pdf"),
    error = TRUE,
    highlight = TRUE,
    message = FALSE,
    prompt = FALSE,
    tidy = FALSE,
    warning = FALSE)

axis <- ggh4x::guide_axis_truncated(
  trunc_lower = unit(0, "npc"),
  trunc_upper = unit(2, "cm")
)
```

# Project details

``` {r}
set.seed(2023)

name_project <- "Co-expression of genes: pbmc_multimodal"
name_pi <- "Merck"

name_file <- gsub(" ", "_", name_project)
name_file <- gsub("-", "_", name_file)
```


<b>Dataset background</b>
A brief overview of what the dataset looks like:

```{r}
# Load dataset
# remotes::install_github("mojaveazure/seurat-disk", quiet=F)
library(SeuratDisk)
seurat <- LoadH5Seurat("../data/pbmc_multimodal.h5seurat")
```

```{r}
# This dataset only came with ADT and SCT assays
# The raw counts were stored in the "counts" slot of the SCT assay
# Creating another RNA assay with the raw counts as CSCORE requires it
raw_rna <- GetAssayData(object =  seurat[['SCT']], layer = 'counts')
seurat[["RNA"]] <- CreateAssayObject(counts = raw_rna )
seurat <- NormalizeData(seurat, assay="RNA")

DefaultAssay(seurat) <- "RNA"
```

```{r fig.height=6}
seurat$celltype <- seurat$celltype.l1
seurat@meta.data %>%
        ggplot() +
        geom_bar(aes(
            x = celltype,
            fill=celltype),
            stat = "count", color = "black") +
        theme_classic() +
        NoLegend() +
        xlab("Celltype") +
        ylab("Number of Cells") +
        ggtitle(glue("{name_project}\nCelltypes")) +
        theme(plot.title = element_text(hjust = 0.5)) +
        geom_text(aes(x = celltype, label = after_stat(count)), stat = "count", vjust = -0.5) +
        theme(axis.text.x = element_text(angle = 45, hjust=1))
```

```{r fig.height=5, fig.width=7}
Idents(seurat) <- "celltype"
p <- DimPlot(seurat, reduction = "umap") +
    xlab("UMAP1") +
    ylab("UMAP2") +
    ggtitle(glue("{name_project}\nCelltypes")) +
    theme(plot.title = element_text(hjust = 0.5)) +
    theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(),
        axis.text.y=element_blank(), axis.ticks.y=element_blank(),
        axis.title=element_text(size=8, face="bold")) +
    guides(x = axis, y = axis) +
    theme(axis.line = element_line(arrow = arrow(type = "closed"), size = 1),
        axis.title = element_text(hjust = 0))
LabelClusters(p, id = "ident",  fontface = "bold", size = 3, bg.colour = "white", bg.r = .2, force = 0)
```

# Protein -> RNA naming

I am going to investigate the dropout rate more closely. 

Since there are multiple different names for a gene, I am scraping the database for alternative names for the various proteins that were used in this dataset so we can look at as many instances of dropout as possible.

I used the resources titled [Data from the Human Protein Atlas in tab-separated format](https://www.proteinatlas.org/about/download) from the Human Protein Atlas to get all possible naming conventions.

```{r}
genes_adt <- sort(row.names(seurat[["ADT"]]))
genes_rna <- sort(row.names(seurat[["RNA"]]))
```

```{r}
df <- read.csv("~/Downloads/proteinatlas.tsv", row.names=NULL, sep="\t")
df <- df[c("Gene", "Gene.synonym")]

list_gene <- c()
list_syno <- c()

for (idx in 1:nrow(df)) {
    gene <- df[idx, "Gene"]
    gene.synonym <- df[idx, "Gene.synonym"]
    gene.synonym <- trimws(strsplit(gene.synonym, ",")[[1]])

    for (g in gene.synonym) {
        list_gene <- c(list_gene, gene)
        list_syno <- c(list_syno, g)
    }
}

df_gene <- data.frame(gene=list_gene, gene_syn=list_syno)
df_gene <- df_gene %>% mutate(across(where(is.character), toupper))

genes <- genes_adt[(genes_adt %in% df_gene$gene) | (genes_adt %in% df_gene$gene_syn)]

list_protein <- c()
list_rna <- c()
for (protein in genes) {
    df_protein <- df_gene[(df_gene$gene == protein) | (df_gene$gene_syn == protein),]
    gene_names <- unlist(unname(as.list(df_protein)))

    rna <- unique(gene_names[gene_names %in% genes_rna])
    if (rlang::is_empty(rna) | length(rna) > 1) {
        print(paste0(protein, " : ", rna))
        next
    }

    list_protein <- c(list_protein, protein)
    list_rna <- c(list_rna, rna)
}


df_genes <- data.frame(protein=list_protein, rna=list_rna) 
df_genes
```

# RNA vs Protein expression

Here I am applying a log10 tranformation to the normalized protein and RNA matrices for the genes that we are interested in.

```{r}
# Log scale the ADT counts
counts_adt <- FetchData(seurat[["ADT"]], vars=df_genes$protein)
counts_adt <- log10(counts_adt + 0.01) - log10(0.01)

# Log scale the RNA normalized counts
counts_rna <- FetchData(seurat[["RNA"]], vars=df_genes$rna)
counts_rna <- log10(counts_rna + 0.01) - log10(0.01)

# Celltype information for each cell
celltype <- unname(seurat$celltype)
counts_rna$celltype <- celltype
counts_adt$celltype <- celltype
```

# Calculate: %dropout + correlation

Here I am subsetting the cells by celltype. Then using the list of genes (shown above) I am computing the proportion of dropout (RNA expression == 0) as well as the pearson correlation coefficient when comparing RNA vs Protein expression.

```{r}
seurat$celltype <- seurat$celltype.l1
celltypes <- sort(unique(seurat$celltype))

list_dropout <- list()
list_corr <- list()

for (ct in celltypes) {
    print(ct)

    counts_adt_ct <- counts_adt[counts_adt$celltype == ct, ]
    counts_rna_ct <- counts_rna[counts_rna$celltype == ct, ]

    ct_dropout <- c()
    ct_corr <- c()

    for (idx in 1:nrow(df_genes)) {
        protein <- df_genes[idx, "protein"]
        gene <- df_genes[idx, "rna"]

        pct_dropout <- sum(counts_rna_ct[gene] == 0.0) / dim(counts_rna_ct)[1]
        ct_dropout <- c(ct_dropout, pct_dropout)

        # Subset out 0 RNA expression
        df <- data.frame(counts_rna_ct[[gene]], counts_adt_ct[[protein]])
        colnames(df) <- c("RNA", "Protein")
        df <- df[df$RNA != 0,]

        correlation <- cor(counts_rna_ct[[gene]], counts_adt_ct[[protein]], method="pearson") %>% as.numeric
        ct_corr <- c(ct_corr, correlation)
    }

    names(ct_dropout) <- df_genes$rna
    names(ct_corr) <- df_genes$rna

    list_dropout[[ct]] <- ct_dropout
    list_corr[[ct]] <- ct_corr
}

df_dropout <- data.frame(list_dropout)
df_corr <- data.frame(list_corr)
```

## % dropout

Percentage of instances where a cell has RNA expression = 0 for a particular gene within a celltype.

```{r}
df_dropout
```

```{r}
df_dropout_melt <- df_dropout %>%
                    rownames_to_column("gene") %>%
                    pivot_longer(cols = -gene, names_to = "celltype", values_to = 'pct_dropout')

ggplot(df_dropout_melt, aes(pct_dropout)) +
    geom_histogram(bins=50, alpha = 0.3, position = "identity", fill="pink", color="pink") +
    theme_classic() +
    xlab("% dropout (RNA expression == 0)") +
    ggtitle("% dropout per celltype/gene")
```

## RNA vs protein correlation

When computing these correlation values, I first subset the count matrix by celltype. Then, I removed the cells that had 0 expression for the corresponding gene in the RNA. I then computed the pearson correlation on the remaining values, comparing rna vs protein expression.

```{r}
df_corr
```

```{r}
df_corr_melt <- df_corr %>%
                    rownames_to_column("gene") %>%
                    pivot_longer(cols = -gene, names_to = "celltype", values_to = 'corr')

ggplot(df_corr_melt, aes(corr)) +
    geom_histogram(bins=50, alpha = 0.3, position = "identity", fill="purple", color="purple") +
    theme_classic() +
    xlab("Pearson correlation coefficient") +
    ggtitle("Pearson correlation: RNA vs protein")
```


## Dropout vs correlation

```{r}
df <- merge(df_corr_melt, df_dropout_melt, by=c("celltype", "gene"))

df
```

```{r}
ggplot(df, aes(x=corr, y=pct_dropout, color=celltype)) +
    geom_point() +
    theme_classic() +
    xlab("Correlation") +
    ylab("Dropout")
```

# Remove %dropout > 0.95

```{r}
df_sub <- df[df$pct_dropout < 0.95, ]
```

```{r}
ggplot(df_sub, aes(x=corr, y=pct_dropout, color=celltype)) +
    geom_point() +
    theme_classic() +
    xlab("Correlation") +
    ylab("Dropout")
```

```{r}
ggplot(df_sub, aes(pct_dropout)) +
    geom_histogram(bins=50, alpha = 0.3, position = "identity", fill="pink", color="pink") +
    theme_classic() +
    xlab("% dropout (RNA expression == 0)") +
    ggtitle("% dropout per celltype/gene")
```

```{r}
ggplot(df_sub, aes(corr)) +
    geom_histogram(bins=50, alpha = 0.3, position = "identity", fill="purple", color="purple") +
    theme_classic() +
    xlab("Pearson correlation coefficient") +
    ggtitle("Pearson correlation: RNA vs protein")
```

# Remove corr != 0

```{r}
df_sub <- df[!(is.na(df$corr)),]
df_sub <- df_sub[(df_sub$corr > 0.05) | (df_sub$corr < -0.05), ]
```

```{r}
ggplot(df_sub, aes(x=corr, y=pct_dropout, color=celltype)) +
    geom_point() +
    theme_classic() +
    xlab("Correlation") +
    ylab("Dropout")
```

```{r}
ggplot(df_sub, aes(pct_dropout)) +
    geom_histogram(bins=50, alpha = 0.3, position = "identity", fill="pink", color="pink") +
    theme_classic() +
    xlab("% dropout (RNA expression == 0)") +
    ggtitle("% dropout per celltype/gene")
```

```{r}
ggplot(df_sub, aes(corr)) +
    geom_histogram(bins=50, alpha = 0.3, position = "identity", fill="purple", color="purple") +
    theme_classic() +
    xlab("Pearson correlation coefficient") +
    ggtitle("Pearson correlation: RNA vs protein")
```