---
title: "Protein expression for pbmc_multimodal"
author: "Noor Sohail"
output:
   html_document:
      code_folding: hide
      df_print: paged
      highlights: pygments
      number_sections: true
      self_contained: true
      theme: default
      toc: true
      toc_float:
         collapsed: true
         smooth_scroll: true
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
# Turn off Warnings and other console output messages from the whole document
```

```{r, cache=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(glue)
library(Seurat)
library(scales)
library(ggh4x)
library(gridExtra)
library(pheatmap)
library(devtools)

library(CSCORE)
library(WGCNA)
library(ggpubr)
library(ggvenn)

library(SAVER)

ggplot2::theme_set(theme_light(base_size = 11))
opts_chunk[["set"]](
    cache = FALSE,
    dev = c("png", "pdf"),
    error = TRUE,
    highlight = TRUE,
    message = FALSE,
    prompt = FALSE,
    tidy = FALSE,
    warning = FALSE)

axis <- ggh4x::guide_axis_truncated(
  trunc_lower = unit(0, "npc"),
  trunc_upper = unit(2, "cm")
)
```

# Project details

``` {r}
set.seed(2023)

name_project <- "Co-expression of genes: pbmc_multimodal"
name_pi <- "Merck"

name_file <- gsub(" ", "_", name_project)
name_file <- gsub("-", "_", name_file)
```


<b>Dataset background</b>
A brief overview of what the dataset looks like:

```{r}
# Load dataset
# remotes::install_github("mojaveazure/seurat-disk", quiet=F)
library(SeuratDisk)
seurat <- LoadH5Seurat("../data/pbmc_multimodal.h5seurat")
```

```{r}
# This dataset only came with ADT and SCT assays
# The raw counts were stored in the "counts" slot of the SCT assay
# Creating another RNA assay with the raw counts as CSCORE requires it
raw_rna <- GetAssayData(object =  seurat[['SCT']], layer = 'counts')
seurat[["RNA"]] <- CreateAssayObject(counts = raw_rna )
seurat <- NormalizeData(seurat, assay="RNA")

DefaultAssay(seurat) <- "RNA"
```

```{r fig.height=6}
seurat$celltype <- seurat$celltype.l1
seurat@meta.data %>%
        ggplot() +
        geom_bar(aes(
            x = celltype,
            fill=celltype),
            stat = "count", color = "black") +
        theme_classic() +
        NoLegend() +
        xlab("Celltype") +
        ylab("Number of Cells") +
        ggtitle(glue("{name_project}\nCelltypes")) +
        theme(plot.title = element_text(hjust = 0.5)) +
        geom_text(aes(x = celltype, label = after_stat(count)), stat = "count", vjust = -0.5) +
        theme(axis.text.x = element_text(angle = 45, hjust=1))
```

```{r fig.height=5, fig.width=7}
Idents(seurat) <- "celltype"
p <- DimPlot(seurat, reduction = "umap") +
    xlab("UMAP1") +
    ylab("UMAP2") +
    ggtitle(glue("{name_project}\nCelltypes")) +
    theme(plot.title = element_text(hjust = 0.5)) +
    theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(),
        axis.text.y=element_blank(), axis.ticks.y=element_blank(),
        axis.title=element_text(size=8, face="bold")) +
    guides(x = axis, y = axis) +
    theme(axis.line = element_line(arrow = arrow(type = "closed"), size = 1),
        axis.title = element_text(hjust = 0))
LabelClusters(p, id = "ident",  fontface = "bold", size = 3, bg.colour = "white", bg.r = .2, force = 0)
```

# Surfaceome genes
```{r}
# surfaceome <- read.csv("../data/surfaceome.csv")

# library('biomaRt')
# mart <- useDataset("hsapiens_gene_ensembl", useMart("ensembl"))
# gene_IDs <- getBM(filters= "ensembl_gene_id", attributes= c("ensembl_gene_id","hgnc_symbol"),
#               values = surfaceome$ENSEMBL.Gene.ID, mart= mart)

# surfaceome_genes <- left_join(x = surfaceome, y = gene_IDs, by = c("ENSEMBL.Gene.ID" = "ensembl_gene_id"))
# surfaceome_genes <- surfaceome_genes$hgnc_symbol

# surfaceome_genes <- rownames(seurat$RNA)[rownames(seurat$RNA) %in% surfaceome_genes]

# write.csv(surfaceome_genes, "../data/surfaceome_genes.csv")
surfaceome_genes <- read.csv("../data/surfaceome_genes.csv")$x
as.data.frame(surfaceome_genes)
```

# Subset to CD4 T cells

Since this is just going to be an example of the workflow and things that we have accomplished, I am going to subset the dataset to just the CD4 T cells to simplify this exercise.

```{r}
seurat <- subset(seurat, subset = (celltype == "CD4 T"))

# Removing genes that have 0 counts across all cells of the celltype
counts <- seurat[["RNA"]]@counts
genes.use <- rowSums(counts) > 0
genes.use <- names(genes.use[genes.use])
seurat <- seurat[genes.use, ]

# Get the updated list of surfaceome genes for CD4 cells
surfaceome_genes <- surfaceome_genes[surfaceome_genes %in% row.names(seurat[["RNA"]])]
```

I am also only going to look at a random sample of 15 genes that are the most expressed and found in a majority of the cells.

```{r fig.height=8, fig.width=10}
data_sct <- FetchData(seurat[["SCT"]], vars=surfaceome_genes)

num_cells <- colSums(data_sct > 0)
avg_expression <- colMeans(data_sct)

df_genes <- data.frame(num_cells=num_cells, avg_expression=avg_expression)
df_genes$gene <- row.names(df_genes)

set.seed(2024)
df_genes <- df_genes[with(df_genes, order(avg_expression, num_cells)), ]
surfaceome_genes <- c(sample_n(df_genes, 50)$gene, tail(df_genes, 15)$gene)

ggplot(df_genes, aes(x=num_cells, y=avg_expression)) +
  geom_point() +
  theme_classic() +
  geom_text(aes(label=ifelse(gene %in% surfaceome_genes, as.character(gene), '')), hjust=-0.1, vjust=0.1, color="red")

df_genes[df_genes$gene %in% surfaceome_genes, ]
```

# Imputation

Since dropout is a large problem that we have to contend with in scRNA datasets, we theorized that imputing the matrix to account for this dropout. We have three methods to generate a new, normalized count matrix:

1. SCTransform (raw counts -> normalized counts)
2. MAGIC (raw counts -> normalized counts)
3. SAVER (raw counts -> normalized counts)

```{r}
# Get raw counts
raw_rna <- FetchData(seurat[["RNA"]], vars=surfaceome_genes, layer="counts")
```

## Standard normalization
```{r}
data_rna <- FetchData(seurat[["RNA"]], vars=surfaceome_genes, layer="data")

pheatmap(data_rna, cluster_rows = FALSE, cluster_cols = FALSE, show_rownames = FALSE)
```

## SCTransform

```{r}
data_sct <- data <- FetchData(seurat[["SCT"]], vars=surfaceome_genes, layer="data")

pheatmap(data_sct, cluster_rows = FALSE, cluster_cols = FALSE, show_rownames = FALSE)
```

## MAGIC

```{r}
# MAGIC
# Load conda environment
myenvs <- reticulate::conda_list()
envname <- myenvs$name[3]
reticulate::use_condaenv(envname, required = TRUE)

library(Rmagic)

data_magic <- magic(raw_rna)$result

pheatmap(data_magic, cluster_rows = FALSE, cluster_cols = FALSE, show_rownames = FALSE)
```

## SAVER

```{r}
# Generate SAVER predictions for those genes
data_saver <- saver(t(raw_rna), estimates.only = TRUE, ncores=6)
data_saver <- data.frame(t(data_saver))

pheatmap(data_saver, cluster_rows = FALSE, cluster_cols = FALSE, show_rownames = FALSE)
```

## Average expression

```{r}
df_avg_impute <- data.frame(rna=colMeans(data_rna), sct=colMeans(data_sct), magic=colMeans(data_magic), saver=colMeans(data_saver))
pheatmap(df_avg_impute, cluster_rows = FALSE, cluster_cols = FALSE, show_rownames = TRUE)

df_avg_impute
```

# Correlation

We have several different methods that we can use for compute correlation scores with associated p-values:

1. Pearson correlation 
  - SCTransform counts -> pearson correlation matrix
  - MAGIC imputed -> pearson correlation matrix
  - SAVER imputed -> pearson correlation matrix
2. CS-CORE (raw counts -> co-expression matrix )

## Pearson correlation

### SCT
```{r}
data_sct <- FetchData(seurat[["SCT"]], vars=surfaceome_genes)
pearson_corr <- Hmisc::rcorr(as.matrix(data_sct), type="pearson")

p_corr_sct <- pearson_corr$P
p_corr_sct[is.na(p_corr_sct)] <- 0
corr_sct <- pearson_corr$r
corr_sct[is.na(corr_sct)] <- 1

# pheatmap(corr_sct)
```

### MAGIC
```{r}
pearson_corr <- Hmisc::rcorr(as.matrix(data_magic), type="pearson")

p_corr_magic <- pearson_corr$P
p_corr_magic[is.na(p_corr_magic)] <- 0
corr_magic <- pearson_corr$r
corr_magic[is.na(corr_magic)] <- 1

# pheatmap(corr_magic)
```

### SAVER
```{r}
pearson_corr <- Hmisc::rcorr(as.matrix(data_saver), type="pearson")

p_corr_saver <- pearson_corr$P
p_corr_saver[is.na(p_corr_saver)] <- 0
corr_saver <- pearson_corr$r
corr_saver[is.na(corr_saver)] <- 1

# pheatmap(corr_saver)
```

## CS-CORE

```{r}
CSCORE_result <- CSCORE(seurat, genes = surfaceome_genes)

corr_cscore <- CSCORE_result$est
p_corr_cscore <- CSCORE_result$p_value

# pheatmap(corr_cscore)
```

## CS-CORE vs pearson

Here we want to see if there os agreement between p-value (color) and correlation estimates. 

```{r}
# df_corr <- reshape2::melt(corr_cscore)
# df_corr <- rename(df_corr, c("corr_cscore"="value"))
# df_corr <- merge(df_corr, reshape2::melt(corr_sct))
# df_corr <- rename(df_corr, c("corr_sct"="value"))

# df_corr$p_sct <- reshape2::melt(p_corr_sct)$value < 0.05
# df_corr$p_cscore <- reshape2::melt(p_corr_cscore)$value < 0.05

# df_corr <- df_corr[df_corr$Var1 != df_corr$Var2, ]

# df_corr$significant <- NA
# df_corr[df_corr$p_cscore == TRUE, "significant"] <- "CS-CORE"
# df_corr[df_corr$p_sct == TRUE, "significant"] <- "pearson_sct"
# df_corr[(df_corr$p_sct == TRUE) & (df_corr$p_cscore == TRUE), "significant"] <- "both"
# df_corr[(df_corr$p_sct == FALSE) & (df_corr$p_cscore == FALSE), "significant"] <- "neither"


# pairs <- c()
# for (idx in 1:nrow(df_corr)) {
#     pair <- sort(c(df_corr[idx, ]$Var1, df_corr[idx, ]$Var2))
#     pair <- paste0(pair[1], "+", pair[2])
#     pairs <- c(pairs, c(pair))
# }
# df_corr$pairs <- pairs
# df_corr <- df_corr[!duplicated(df_corr$pairs), ]


# ggplot(df_corr, aes(x=corr_cscore, y=corr_sct, color=significant)) +
#   geom_point() +
#   theme_classic() +
#   geom_abline(slope=1, intercept = 0, color="black") +
#   geom_text(aes(label=pairs), hjust=-0.1, vjust=0.1)
```


# WGCNA modules

Then we can attribute modules to each of the genes. We can supply either the CS-core correlation estimate or the pearson correlation  matrix to see how different the modules appear depending on what we supply to WGCNA.

```{r}
run_wgcna <- function(mtx_coex, mtx_pval, genes) {
  # Obtain BH-adjusted p values
  p_matrix_BH <- matrix(0, length(genes), length(genes))
  p_matrix_BH[upper.tri(p_matrix_BH)] <- p.adjust(mtx_pval[upper.tri(mtx_pval)], method = "BH")
  p_matrix_BH <- p_matrix_BH + t(p_matrix_BH)

  # Set co-expression entires with BH-adjusted p-values greater than 0.05 to 0
  mtx_coex[p_matrix_BH > 0.05] <- 0

  adj <- WGCNA::adjacency.fromSimilarity(abs(mtx_coex), power = 1)
  # Compute the topological overlap matrix
  TOM <- WGCNA::TOMsimilarity(adj)
  dissTOM <- 1 - TOM
  rownames(dissTOM) <- colnames(dissTOM) <- genes
  # Run hierarchical clustering as in the WGCNA workflow
  hclust_dist <- hclust(as.dist(dissTOM), method = "average") 
  memb <- dynamicTreeCut::cutreeDynamic(dendro = hclust_dist, 
                      distM = dissTOM, 
                      deepSplit = 2,
                      pamRespectsDendro = FALSE,
                      minClusterSize = 10)
  # For more instructions on how to tune the parameters in the WGCNA workflow,
  # please refer to https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/Rpackages/WGCNA/Tutorials/

  names(memb) <- genes
  memb_tab <- table(memb)
  module_list <- lapply(sort(unique(memb)), function(i_k) names(which(memb == i_k)))

  module_df <- data.frame(memb)
  module_df$memb <- as.character(module_df$memb)
  df <- module_df
  df <- tibble::rownames_to_column(df, "gene")
  colnames(df) <- c("gene", "module")

  results <- list()
  results[["wgcna"]] <- df
  results[["bh_mtx"]] <- mtx_coex

  return(results)
}
```

## SCT

```{r}
results <- run_wgcna(corr_sct, p_corr_sct, surfaceome_genes)
wgcna_sct <- results$wgcna
mtx_sct <- results$bh_mtx

wgcna_sct
```

```{r}
rownames(wgcna_sct) <- wgcna_sct$gene
anno <- wgcna_sct[, "module", drop=F]
pheatmap(mtx_sct, annotation_row=anno, annotation_col=anno)
```

## MAGIC

```{r}
results <- run_wgcna(corr_magic, p_corr_magic, surfaceome_genes)
wgcna_magic <- results$wgcna
mtx_magic <- results$bh_mtx

wgcna_magic
```

```{r}
rownames(wgcna_magic) <- wgcna_magic$gene
anno <- wgcna_magic[, "module", drop=F]
pheatmap(mtx_magic, annotation_row=anno, annotation_col=anno)
```

## SAVER

```{r}
results <- run_wgcna(corr_saver, p_corr_saver, surfaceome_genes)
wgcna_saver <- results$wgcna
mtx_saver <- results$bh_mtx

wgcna_saver
```

```{r}
rownames(wgcna_saver) <- wgcna_saver$gene
anno <- wgcna_saver[, "module", drop=F]

rownames(mtx_saver) <- str_replace(rownames(mtx_saver), "\\.", "-")
colnames(mtx_saver) <- str_replace(colnames(mtx_saver), "\\.", "-")

pheatmap(mtx_saver, annotation_row=anno, annotation_col=anno)
```

## CS-CORE

```{r}
results <- run_wgcna(corr_cscore, p_corr_cscore, surfaceome_genes)
wgcna_cscore <- results$wgcna
mtx_cscore <- results$bh_mtx

wgcna_cscore
```

```{r}
rownames(wgcna_cscore) <- wgcna_cscore$gene
anno <- wgcna_cscore[, "module", drop=F]
pheatmap(mtx_cscore, annotation_row=anno, annotation_col=anno)
```