---
title: "Correlation workflow"
author: "Noor Sohail"
output:
   html_document:
      code_folding: hide
      df_print: paged
      highlights: pygments
      number_sections: true
      self_contained: true
      theme: default
      toc: true
      toc_float:
         collapsed: true
         smooth_scroll: true
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
# Turn off Warnings and other console output messages from the whole document
```

```{r, cache=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(glue)
library(Seurat)
library(pheatmap)
library(gridExtra)

ggplot2::theme_set(theme_light(base_size = 11))
opts_chunk[["set"]](
    cache = FALSE,
    dev = c("png", "pdf"),
    error = TRUE,
    highlight = TRUE,
    message = FALSE,
    prompt = FALSE,
    tidy = FALSE,
    warning = FALSE)
```

# User define values

```{r}
celltypes <- c("Mono", "CD4 T", "CD8 T", "NK", "B", "other T", "DC")
idx <- 3
```

```{r}
path_seurat <- "../data/pbmc_multimodal.h5seurat"
path_outs <- "results/rm_dropout_2"

col_sample <- "orig.ident"
col_celltype <- "celltype.l1"
ct <- celltypes[idx]

# Filtration parameters
filter <- FALSE
min_exp <- 0.2
min_cells <- 40
min_perc <- 0.2
```

- Path seurat: `r path_seurat`
- Metadata column with celltypes: `r col_celltype`
- Celltype to subset to: `r ct`
- Filter genes based on expression and frequency: `r filter`


# Load seurat object

```{r}
filename <- glue("data/{ct}.RDS")
seurat <- readRDS(filename)
seurat
```


# Genes

List of all genes:

```{r}
# Fill in the list of genes you are interested in calculating correlations for
corr_genes_all <- read.csv("protein_gene_map.csv")$gene
as.data.frame(corr_genes_all)
```

Starting off with `r length(corr_genes_all)` genes of interest.


# Subset to `r ct` cells

```{r}
seurat <- subset(seurat, celltype == ct)

# Manual change
# This dataset only came with ADT and SCT assays
# The raw counts were stored in the "counts" slot of the SCT assay
# Creating another RNA assay with the raw counts as CSCORE requires it
raw_rna <- GetAssayData(object =  seurat[['SCT']], layer = 'counts')
seurat[["RNA"]] <- CreateAssayObject(counts = raw_rna)
seurat <- NormalizeData(seurat, assay = "RNA")

DefaultAssay(seurat) <- "RNA"
# Removing genes that have 0 counts across all cells of the celltype
genes.use <- rowSums(raw_rna) > 0
genes.use <- names(genes.use[genes.use])
seurat <- seurat[genes.use, ]

n_cells <- ncol(seurat)
```

Working with `r n_cells` cells.

```{r}
seurat@meta.data %>%
        ggplot() +
        geom_bar(aes(
            x = get(col_sample),
            fill = get(col_sample)),
            stat = "count", color = "black") +
        theme_classic() +
        NoLegend() +
        ggtitle(glue("{ct} cells: Sample distribution")) +
        ylab("Number of Cells") +
        xlab("Sample") +
        theme(plot.title = element_text(hjust = 0.5)) +
        geom_text(aes(x = get(col_sample), label = after_stat(count)), stat = "count", vjust = -0.5) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Subset genes of interest

From the genes that were supplied, removing any that are not expressed in this dataset.

```{r}
corr_genes <- corr_genes_all[corr_genes_all %in% rownames(seurat)]
as.data.frame(corr_genes)
```

Next we look at the basic distribution of remaining genes of interest in terms of expression and number of cells they are expressed in.

These are the following filtration metrics that are set:

- Average expression < `r min_exp`
- Number of cells expressed in > `r min_cells`
- Percentage of cells expressed in > `r min_perc`

Filtration parameter was set to `r filter`. If FALSE, no further filtration will be done at this step.

```{r}
data_rna <- FetchData(seurat[["RNA"]], vars=corr_genes, layer="data")

# Number of cells a gene is expressed in
num_cells <- colSums(data_rna > 0)
# Percentage of cells a gene is expressed in
perc_cells <- num_cells / ncol(seurat)
# Average expression of a gene
avg_expression <- colMeans(data_rna)

df_genes <- data.frame(num_cells, perc_cells, avg_expression)
df_genes <- df_genes %>% mutate(filter = !((perc_cells > min_perc) & (num_cells > min_cells) & (avg_expression > min_exp)))
df_genes$gene <- row.names(df_genes)
df_genes <- df_genes %>% arrange(desc(avg_expression), desc(perc_cells))

df_genes
```

```{r}
df_genes %>% ggplot() +
    geom_point(aes(x = perc_cells, y = avg_expression, color = filter)) +
    theme_classic()
```

```{r}
if (filter == TRUE) {
    corr_genes <- (df_genes %>% subset(filter == FALSE))$gene
}
```

`r length(corr_genes)` genes of interest remaining.

# Correlation Estimates

Removing cells that have 0 expression of at least one of the genes in the calculation:

```{r}
# Store output so we don't have to re-run correlation each time
filename <- glue("{path_outs}/corr_{ct}.csv")

if (!file.exists(filename)) {

    # Compute spearman correlation for each method (except CS-CORE which is run later)
    # Unique combination of each gene pair
    genes_comb <- data.frame(t(combn(corr_genes, 2)))
    n_comb <- nrow(genes_comb)

    # Create dataframe with correlation and p-values scores
    df_corr <- genes_comb %>% rename("Var1" = X1, "Var2" = X2)
    df_corr["rm_dropout"] <- NA
    df_p_val <- df_corr

    for (idx in 1:n_comb) {
        assay_ <- "rm_dropout"

        if (idx %% 200 == 0) {
            print(glue("{idx}/{n_comb} correlations computed."))
        }

        # Name of genes to run correlation on
        gene_1 <- genes_comb[idx, 1]
        gene_2 <- genes_comb[idx, 2]

        gene_exp <- FetchData(seurat, assay="RNA", layer="data", vars=c(gene_1, gene_2))
        gene_exp <- gene_exp %>% subset((get(gene_1) > 0) & (get(gene_2) > 0))


        if (nrow(gene_exp) <= 1) {
            # If a gene has no expression, set correlation = 0 and p-value = 1
            corr_val <- 0.0
            p_val <- 1.0
        } else {
            # Calculate spearman correlation and p-value otherwise
            tmp <- cor.test(gene_exp[[gene_1]], gene_exp[[gene_2]], method = "spearman", exact = FALSE)
            corr_val <- as.numeric(unname(tmp$estimate))
            p_val <- as.numeric(tmp$p.value)
        }

        # Store correlation and p-values
        df_corr[idx, assay_] <- corr_val
        df_p_val[idx, assay_] <- p_val
    }

    # Save output
    write.csv(df_corr, filename)
    write.csv(df_p_val, glue("{path_outs}/p_corr_{ct}.csv"))
}

df_corr <- read.csv(filename, row.names=1)
df_p_val <- read.csv(glue("{path_outs}/p_corr_{ct}.csv"), row.names=1)
```

## Heatmap of correlation estimates

Showing the patterns of correlation for each method. The x-axis and y-axis are the genes of interest with the corresponding correlation value for the pair as the value. Keep in mind that this is symmetric matrix.

```{r results = "asis"}
library(RColorBrewer)
methods <- c("rm_dropout")

for (method in methods) {
    corr <- df_corr[c("Var1", "Var2", method)]
    corr_cp <- corr %>% rename(Var1 = Var2, Var2 = Var1)
    corr <- rbind(corr, corr_cp)
    mtx <- reshape2::dcast(corr, Var2 ~ Var1) %>% column_to_rownames("Var2")

    # Set the diagonal values: Correlation = 1, p-value = 1
    mtx <- as.matrix(mtx)
    diag(mtx) <- 1

    breaks <-  seq(-1, 1, by = 0.1)
    color <- colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(length(breaks))


    p <- pheatmap(mtx, show_rownames = FALSE, show_colnames = FALSE,
                color = color, breaks = breaks, silent = FALSE, main = method)
    knit_print(p)
}
```
