---
title: "DC: Sequencing depth effect on correlations"
author: "Noor Sohail"
output:
   html_document:
      code_folding: hide
      df_print: paged
      highlights: pygments
      number_sections: true
      self_contained: true
      theme: default
      toc: true
      toc_float:
         collapsed: true
         smooth_scroll: true
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
# Turn off Warnings and other console output messages from the whole document
```

```{r}
library(reticulate)
use_condaenv("merck")
```

```{r, cache=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(glue)
library(Seurat)
library(pheatmap)
library(devtools)
library(gridExtra)
library(RColorBrewer)

library(ggpubr)

ggplot2::theme_set(theme_light(base_size = 11))
opts_chunk[["set"]](
    cache = FALSE,
    dev = c("png", "pdf"),
    error = TRUE,
    highlight = TRUE,
    message = FALSE,
    prompt = FALSE,
    tidy = FALSE,
    warning = FALSE)
```

# Objective

Group cells together by their library size/sequencing depth and re-calculate correlations for gene pairs within those groups. If sequencing depth has a large effect on correlation scores, we would expect to see large differences when comparing low library size correlations against higher sequencing depth correlations.

This report look at the DC cells from the PBMC multimodal (RNA + ADT) dataset.

# User define values


```{r}
path_seurat <- "data/DC.RDS"
path_outs <- "results/"

## add column names for the sample id and the celltype column. ct is the cluster/celltype that you want to focus on. Replace with your information that match seurat object.
col_sample <- "donor"
col_celltype <- "celltype"
ct <- "DC"

# Filtration parameters
filter <- FALSE
min_exp <- 0.2
min_cells <- 40
min_perc <- 0.2
```

- Path seurat: `r path_seurat`
- Metadata column with celltypes: `r col_celltype`
- Celltype to subset to: `r ct`
- Filter genes based on expression and frequency: `r filter`

To help evaluate the results, I am using a set of control genes that are known to be expressed in DC cells:

```{r}
# Fill in the list of genes you are interested in calculating correlations for
corr_genes_all <- c("CD1C", "FCER1A", "CLEC10A", "PTPRC", "IRF4", "CEBPB", "LILRB4A", "ITGAX", "FSCN1", "CCR7")

as.data.frame(corr_genes_all)
```


# Load seurat object

```{r}
seurat <- readRDS(path_seurat)

raw_rna <- GetAssayData(object =  seurat[['SCT']], layer = 'counts')
seurat[["RNA"]] <- CreateAssayObject(counts = raw_rna)
seurat <- NormalizeData(seurat, assay = "RNA")

# Saving celltype information as column named celltype
# In order to use subset function later
seurat$celltype <- seurat@meta.data[col_celltype]
```

```{r}
## Set default assay to RNA, downstream code fails if it is on SCT. Remove SCT as well
DefaultAssay(object = seurat) <- "RNA"
# seurat[['SCT']] <- NULL # to make sure it doesn't fail downstream due to new version of Seurat
n_cells <- ncol(seurat)
```

Working with `r n_cells` cells.

```{r}
seurat@meta.data %>%
        ggplot() +
        geom_bar(aes(
            x = get(col_sample),
            fill = get(col_sample)),
            stat = "count", color = "black") +
        theme_classic() +
        NoLegend() +
        ggtitle(glue("{ct} cells: Sample distribution")) +
        ylab("Number of Cells") +
        xlab("Sample") +
        theme(plot.title = element_text(hjust = 0.5)) +
        geom_text(aes(x = get(col_sample), label = after_stat(count)), stat = "count", vjust = -0.5) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Selecting extra genes

The genes selected so far are DC marker genes and are being used as positive controls for this analysis.

Next I am going to select some genes based upon the average expression, selecting genes in each group to get a good range of expression values

```{r}
gene_avg <- rowMeans(seurat[["RNA"]]$data) %>% as.data.frame() %>% rename("avg"=".")
qts <- quantile(gene_avg$avg, probs=seq(0, 1, 0.1))

genes <- c()
for (idx in 1:10) {
  group <- glue("group_{idx}")
  
  qt1 <- qts[idx]
  qt2 <- qts[idx + 1]
  
  gene <- gene_avg %>% subset((avg >= qt1) & (avg <= qt2)) %>% rownames()
  genes <- c(genes, gene[100])
}

gene_avg %>% rownames_to_column("gene") %>% subset(gene %in% genes)
```


# Subset genes of interest

From the genes that were supplied, removing any that are not expressed in this dataset.

Here we can see the final list of genes that are used in the analysis: 

```{r}
corr_genes_all <- c(corr_genes_all, genes)

corr_genes_all <- c("CD1C", "FCER1A", "CLEC10A", "GPX1", "PTPRC", "IRF4", "CEBPB", "LILRB4A", "ITGAX", "FSCN1", "CCR7", "SMIM10L2A", "PTPRN", "AL451074.2", "CHI3L2", "KCNA3", "AC239800.3", "ABCD3", "CEPT1", "MCOLN2", "PRPF38B")

corr_genes <- corr_genes_all[corr_genes_all %in% rownames(seurat)]
as.data.frame(corr_genes)
```

Next we look at the basic distribution of remaining genes of interest in terms of expression and number of cells they are expressed in.

These are the following filtration metrics that are set:

- Average expression < `r min_exp`
- Number of cells expressed in > `r min_cells`
- Percentage of cells expressed in > `r min_perc`

Filtration parameter was set to `r filter`. If FALSE, no further filtration will be done at this step.

```{r}
data_rna <- FetchData(seurat[["RNA"]], vars = corr_genes)

# Number of cells a gene is expressed in
num_cells <- colSums(data_rna > 0)
# Percentage of cells a gene is expressed in
perc_cells <- num_cells / ncol(seurat)
# Average expression of a gene
avg_expression <- colMeans(data_rna)

df_genes <- data.frame(num_cells, perc_cells, avg_expression)
df_genes <- df_genes %>% mutate(filter = !((perc_cells > min_perc) & (num_cells > min_cells) & (avg_expression > min_exp)))
df_genes$gene <- row.names(df_genes)
df_genes <- df_genes %>% arrange(desc(avg_expression), desc(perc_cells))

df_genes
```

```{r}
df_genes %>% ggplot() +
    geom_point(aes(x = perc_cells, y = avg_expression, color = filter)) +
    theme_classic() +
    labs(x="Percent cells expressed", y="Average expression")
```

```{r}
if (filter == TRUE) {
    corr_genes <- (df_genes %>% subset(filter == FALSE))$gene
}
```

`r length(corr_genes)` genes of interest remaining.

# Group cells by sequencing depth

Next, we are grouping together cells based upon their total expression value to then calculate correlations among those groups. Here, I am printing out the different quantile values used to make 10 groups of cells.

```{r}
groups <- c("group_1", "group_2", "group_3", "group_4", "group_5", "group_6", "group_7", "group_8", "group_9",  "group_10")
groups <- factor(groups, levels=groups)

qts <- quantile(seurat@meta.data$nCount_RNA, probs=seq(0, 1, 0.1))
qts
```

```{r}
seurat$depth_group <- NA
for (idx in 1:10) {
  group <- glue("group_{idx}")
  
  qt1 <- qts[idx]
  qt2 <- qts[idx + 1]
  
  idx_cells <- seurat@meta.data %>% subset((nCount_RNA >= qt1) & (nCount_RNA <= qt2)) %>% rownames()
  seurat@meta.data[idx_cells, "depth_group"] <- group
}

seurat$depth_group <- factor(seurat$depth_group, levels=groups)
```

This histogram represnts the distribution of sequencing depth for the entire dataset, while also showing the groups that were assigned based upon quantile scores.

```{r}
ggplot(seurat@meta.data) +
  geom_histogram(aes(x=nCount_RNA, fill=depth_group), bins=100) +
  scale_x_log10() +
  theme_classic()
```

We also want to make sure that we have a good number of cells in each group before running the correlation calculation.

```{r}
seurat@meta.data %>%
        ggplot() +
        geom_bar(aes(
            x = depth_group,
            fill = depth_group),
            stat = "count", color = "black") +
        theme_classic() +
        NoLegend() +
        ggtitle(glue("{ct} cells")) +
        ylab("Number of Cells") +
        xlab("Sample") +
        theme(plot.title = element_text(hjust = 0.5)) +
        geom_text(aes(x = depth_group, label = after_stat(count)), stat = "count", vjust = -0.5) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Lastly, we want to take quick look at the expression of our genes across each of the groups we have just defined to make better sense of the results.

```{r fig.width=12, fig.height=8}
Idents(seurat) <- "depth_group"
DotPlot(seurat, corr_genes_all) + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

# Calculate correlations within groups

Here I load in the seurat output `imputed_{celltype}.rds` from the correlation workflow script to re-calculate correlations on each of the subsets.

```{r}
depth <- seurat@meta.data[c("nCount_RNA", "depth_group")]

seurat <- readRDS("data/imputed_DC.RDS")
seurat$nCount_RNA <- depth$nCount_RNA
seurat$depth_group <- depth$depth_group
```

Iterating over each cell group and calculating the spearman correlation on each normalization/imputation method (RNA, SCT, MAGIC, SAVER).

```{r}
# # Compute spearman correlation for each method (except CS-CORE which is run later)
# # Unique combination of each gene pair
# genes_comb <- data.frame(t(combn(corr_genes, 2)))
# n_comb <- nrow(genes_comb)
# 
# df_corr_list <- list()
# df_p_val_list <- list()
# 
# assays <- c("RNA", "SCT", "MAGIC", "SAVER")
# for (assay_ in assays) {
# 
#   # Create dataframe with correlation and p-values scores
#   df_corr <- genes_comb %>% rename("Var1" = X1, "Var2" = X2)
#   df_corr[levels(groups)] <- NA
#   df_p_val <- df_corr
# 
#   for (idx in 1:n_comb) {
# 
#       if (idx %% 100 == 0) {
#           print(glue("{assay_}: {idx}/{n_comb} correlations computed."))
#       }
# 
#       # Name of genes to run correlation on
#       gene_1 <- genes_comb[idx, 1]
#       gene_2 <- genes_comb[idx, 2]
# 
#       for (group_ in groups) {
#         seurat_sub <- subset(seurat, subset = (depth_group == group_))
#         gene_exp <- t(seurat_sub[[assay_]]$data[c(gene_1, gene_2), ]) %>% as.data.frame()
# 
#           if (all(gene_exp[[gene_1]] == 0) | all(gene_exp[[gene_2]] == 0)) {
#               # If a gene has no expression, set correlation = 0 and p-value = 1
#               corr_val <- 0.0
#               p_val <- 1.0
#           } else {
#               # Calculate spearman correlation and p-value otherwise
#               tmp <- cor.test(gene_exp[[gene_1]], gene_exp[[gene_2]],
#                               method = "spearman", exact = FALSE)
#               corr_val <- as.numeric(unname(tmp$estimate))
#               p_val <- as.numeric(tmp$p.value)
#           }
# 
#           # Store correlation and p-values
#           df_corr[idx, group_] <- corr_val
#           df_p_val[idx, group_] <- p_val
#       }
#   }
# 
#   df_corr_list[[assay_]] <- df_corr
#   df_p_val_list[[assay_]] <- df_p_val
# 
# }
# 
# saveRDS(df_corr_list, "results/DC_seqDepth_subset.RDS")
```

Calculating CS-CORE co-expression estimates on each sequencing depth group.

```{r}
# # Create dataframe with correlation and p-values scores
# df_corr <- genes_comb %>% rename("Var1" = X1, "Var2" = X2)
# df_p_val <- df_corr
# 
# for (group_ in groups) {
#   seurat_sub <- subset(seurat, subset = (depth_group == group_))
# 
#   # Run CS-CORE
#   DefaultAssay(seurat_sub) <- "RNA"
#   CSCORE_result <- CSCORE(seurat_sub, genes=corr_genes)
# 
#   # Store CS-CORE results
#   tmp <- reshape2::melt(as.matrix(CSCORE_result$est))
#   idx_col <- which(colnames(tmp) == "value")
#   colnames(tmp)[idx_col] <- group_
# 
#   df_corr <- left_join(df_corr, tmp)
# }
# 
# df_corr_list[["CSCORE"]] <- df_corr
# saveRDS(df_corr_list, "results/DC_seqDepth_subset.RDS")
```

# Results

Here we evaluate the differences in correlation scores when cells are grouped together based upon their sequencing depth.

```{r}
df_corr_list <- readRDS("results/DC_seqDepth_subset.RDS")
```

```{r results="asis"}
methods <- names(df_corr_list)

for (method in methods) {
  cat(paste("##", method, "{.tabset} \n\n"))
  
  cat("### Heatmap \n\n")
  cat("We evaluate the change in correlation coefficients across each group, represented as a heatmap, where the rows represent all gene pairs.")
  cat("\n\n")
  df <- df_corr_list[[method]] %>% select(groups)
  if (method == "SAVER") {
    p <- pheatmap(df, cluster_cols=FALSE, cluster_rows=FALSE, show_rownames = FALSE, main=method, silent=FALSE, na_col = "grey90")
  } else {
    p <- pheatmap(df, cluster_cols=FALSE, show_rownames = FALSE, main=method, silent=FALSE, na_col = "grey90")
  }
  knit_print(p)
  cat("\n\n")

  cat("### Standard deviation distribution\n\n")
  cat("To better understand how much the coefficients vary across the different sequencing depth groups, I visualize the distribution of standard deviations.")
  cat("\n\n")
  df <- transform(df, SD=apply(df,1, sd, na.rm = TRUE))
  p <- ggplot(df, aes(x=SD)) +
    geom_histogram(bins=50) +
    labs(x="Standard deviation", title=method) +
    theme_classic()
  knit_print(p)
  cat("\n\n")

  cat("### Scatterplot group 1 vs 10\n\n")
  cat("I show the differences in correlations between groups 1 and 10. Group 1 contains cells with the lowest library size while 10 has highest library size. Each point represents a gene pair, and is additionally colored by the standard deviation (calculated on all groups).")
  cat("\n\n")
  p <- ggplot(df, aes(x=group_1, y=group_10, color=SD)) +
    geom_point() +
    scale_colour_viridis_c() +
    theme_classic() +
    xlim(-1, 1) + ylim(-1, 1) +
    geom_abline(slope = 1, intercept = 0, color = "red") +
    stat_cor(method="pearson", label.y.npc="top", label.x.npc = "left", color="red") +
    ggtitle(method)
  knit_print(p)
  cat("\n\n")
}
```

# Final thoughts based on results

It is clear that the correlation scores do not differ much across each of the sequencing depth groups. The range of standard deviation values is minimal and when comparing the two expremees (group 1 and group 10), we see that there is very good concordence between the correlation scores across all methods. This leads us to believe that sequencing depth does not have a strong effect on these correlation calcluations. 

